---
title: "Introduction to LDxBlocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LDxBlocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
library(LDxBlocks)
data(ldx_geno,     package = "LDxBlocks")
data(ldx_snp_info, package = "LDxBlocks")
data(ldx_blocks,   package = "LDxBlocks")
data(ldx_gwas,     package = "LDxBlocks")
```

## 1. Overview

`LDxBlocks` detects linkage disequilibrium (LD) blocks from genome-wide SNP
data and extends the pipeline into haplotype reconstruction, diversity
analysis, and genomic prediction feature engineering. Two LD metrics are
available:

- **Standard r²** -- fast, no kinship correction, suitable for 10 M+ markers
  and unstructured populations.
- **Kinship-adjusted rV²** -- corrects for relatedness-induced LD inflation,
  for livestock, inbred lines, or family-based cohorts. See the
  *LD Metrics* vignette for the statistical detail and decision table.

The C++/Armadillo computational core makes the algorithm approximately 40x
faster than the original Big-LD implementation (no compiled code) for typical window
sizes.

This vignette covers the complete pipeline -- reading genotype data, block
detection, haplotype analysis, and parameter tuning -- using the `ldx_geno`
example dataset (120 individuals, 230 SNPs, 3 chromosomes, 9 simulated LD
blocks).

---

## 2. Why LDxBlocks? Relationship to the original Big-LD

LDxBlocks is built on the clique-based segmentation algorithm of Kim et al.
(2018), which introduced interval graph modelling of LD bins as a principled
alternative to sliding-window approaches. The mathematical core -- CLQD bin
assignment, maximum-weight independent set block construction, and
Bron-Kerbosch clique enumeration -- is preserved exactly. LDxBlocks extends
that foundation to address three limitations of the original implementation
that become critical for modern breeding and genomics programmes.

### 2.1 Computational bottleneck

The original `Big_LD()` calls `cor()` inside the boundary-scan loop and once
per `CLQD()` call, with all matrix operations running through the R
interpreter. For a 50,000-SNP chromosome with the default `leng = 200` this
amounts to approximately 150,000 small R-level matrix multiplications per
chromosome. For a WGS rice panel with 3 million SNPs across 12 chromosomes,
the original implementation would take days on a workstation.

LDxBlocks replaces the critical paths with eight compiled C++ functions:

```r
# These calls are now C++/Armadillo + OpenMP -- not R loops
compute_r2_cpp(geno, digits = -1L, n_threads = 8L)   # ~40x faster
maf_filter_cpp(geno, maf_cut = 0.05)                  # ~10x faster
boundary_scan_cpp(geno, start, end, half_w, threshold) # ~20x faster
build_hap_strings_cpp(blk_int, na_char)               # ~20-50x faster
```

The outer loop of `compute_r2_cpp()` is parallelised with OpenMP. For a
1,500-SNP window with 500 individuals, the r² matrix computes in milliseconds
rather than seconds.

### 2.2 Memory wall

The original `Big_LD()` requires the complete genotype matrix in RAM before
detection begins. For a 10M-marker, 5,000-individual WGS dataset this is
approximately 400 GB as an R `double` matrix -- impossible on any workstation.

LDxBlocks enforces a strict never-full-genome memory model. All six supported
formats stream genotypes one chromosome window at a time:

```r
# VCF auto-converts to GDS cache; detection streams one window at a time
be     <- read_geno("wgs_panel.vcf.gz")
blocks <- run_Big_LD_all_chr(be, method = "r2", n_threads = 8L)
# Peak RAM ~ n_samples x subSegmSize x 8 bytes = 60 MB for n=5000, w=1500
```

### 2.3 Pipeline gap

The original Big-LD stops at the block table. For breeding programmes the
immediate next questions are: What haplotypes exist? How diverse are they?
Which blocks harbour QTLs? How do I build a multi-locus genomic prediction
model? LDxBlocks provides a complete answer:

```r
# From blocks to genomic prediction features in four lines
haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)
qtl  <- define_qtl_regions(gwas_results, blocks, be$snp_info)
feat <- build_haplotype_feature_matrix(haps, top_n = 5, encoding = "additive_012")$matrix
```

### 2.4 What is preserved from the original

The following components are unchanged from Kim et al. (2018):

- `CLQD()` (internal): bin vector assignment via maximal clique enumeration and
  greedy density/maximal priority selection
- `constructLDblock()` (internal): maximum-weight independent set via dynamic
  programming on sorted interval sequences
- All `CLQmode`, `clstgap`, and `split` logic
- Block table column format (`start`, `end`, `start.rsID`, `end.rsID`,
  `start.bp`, `end.bp`) -- drop-in compatible with tools that accept original
  Big-LD output

### 2.5 Comparison with related tools

| Feature | Original Big-LD | gpart (Bioconductor) | PLINK --blocks | LDxBlocks |
|---|---|---|---|---|
| Algorithm | Clique/MWIS | Clique/MWIS + GPART | Gabriel et al. | Clique/MWIS |
| LD metric | r | r, D' | r² | r², rV² |
| C++ core | No | Partial | Yes | Yes (9 functions) |
| WGS streaming | No | Partial | Yes | Yes (all formats) |
| Kinship correction | No | No | No | Yes (rV²) |
| Haplotype analysis | No | No | No | Yes |
| Genomic prediction | No | No | No | Yes |
| GWAS-driven tuning | No | No | No | Yes |
| Input formats | R matrix | PLINK, VCF | PLINK | 6 formats |
| Output | Block table | Block table + gene regions | Block table | Block table + 19 downstream functions |

---

## 3. Example data

Four datasets are shipped with the package, all generated by
`data-raw/generate_example_data.R`:

```{r example-data}
dim(ldx_geno)             # 120 individuals x 230 SNPs
head(ldx_snp_info)        # SNP, CHR, POS, REF, ALT
table(ldx_snp_info$CHR)   # chr1, chr2, chr3
nrow(ldx_gwas)            # toy GWAS markers
```

Each chromosome has three LD blocks separated by inter-block gaps and
short stretches of low-LD singleton SNPs.

---

## 4. Step 1 — Reading genotype data

`read_geno()` auto-detects the file format from the extension and returns an
`LDxBlocks_backend` object. Six formats are supported:

| Format | Extension | `format =` |
|--------|-----------|------------|
| Numeric dosage | `.csv`, `.txt` | `"numeric"` |
| HapMap | `.hmp.txt` | `"hapmap"` |
| VCF / bgzipped VCF | `.vcf`, `.vcf.gz` | `"vcf"` |
| SNPRelate GDS | `.gds` | `"gds"` |
| PLINK binary | `.bed` (+ `.bim`, `.fam`) | `"bed"` |
| R matrix | (in-memory) | `"matrix"` |

```{r read-formats, eval=FALSE}
be_vcf <- read_geno("mydata.vcf.gz")
be_csv <- read_geno("mydata.csv")
be_hmp <- read_geno("mydata.hmp.txt")
be_gds <- read_geno("mydata.gds")    # SNPRelate GDS; streams SNP windows
be_bed <- read_geno("mydata.bed")    # PLINK BED; memory-mapped
```

```{r read-matrix}
be <- read_geno(ldx_geno, format = "matrix", snp_info = ldx_snp_info)
be
```

```{r backend-interface}
be$n_samples
be$n_snps
be$sample_ids[1:4]
head(be$snp_info)
chunk <- read_chunk(be, 1:30)
dim(chunk)
close_backend(be)
```


---

## 5. Step 2 — Phenotype input format

The `blues` argument accepts pre-adjusted phenotype means from a mixed model.
The format is flexible — any of four structures are accepted:

**Format 1: Named numeric vector** (single trait, simplest)
```{r blues-vector, eval=FALSE}
blues <- c(ind001 = 4.21, ind002 = 3.87, ind003 = 5.14)
# names = individual IDs matching rownames(geno_matrix)
```

**Format 2: Data frame, single trait**

This is the most common format when reading output from ASReml-R, lme4, or SpATS.
Any column names are accepted — just tell the function which columns to use:

```{r blues-df-single, eval=FALSE}
blues <- read.csv("blues.csv")  # e.g. columns: "Genotype", "YLD_BLUE"
res   <- run_haplotype_prediction(geno, snp_info, blocks,
                                   blues    = blues,
                                   id_col   = "Genotype",   # ID column name
                                   blue_col = "YLD_BLUE")   # BLUE column name
```

**Format 3: Data frame, multiple traits**

```{r blues-df-multi, eval=FALSE}
blues_mt <- read.csv("blues_mt.csv")  # columns: id, YLD, DIS, PHT
res_mt   <- run_haplotype_prediction(geno, snp_info, blocks,
                                      blues           = blues_mt,
                                      id_col          = "id",
                                      blue_cols       = c("YLD", "DIS", "PHT"),
                                      importance_rule = "any")
```

When `blue_cols = NULL` (default), all numeric non-ID columns are used automatically.

**Format 4: Named list** (different individuals per trait)

```{r blues-list, eval=FALSE}
blues <- list(
  YLD = c(ind001 = 4.21, ind002 = 3.87),
  DIS = c(ind001 = 0.32, ind003 = 0.28)  # ind002 missing, ind003 added
)
```

**Column requirements and ID matching**

| Element | Requirement |
|---------|-------------|
| ID column | Any name — set via `id_col`. Must match `rownames(geno_matrix)` exactly (case-sensitive). |
| BLUE column(s) | Any name — set via `blue_col` (single) or `blue_cols` (multi). Must be numeric. |
| Missing values | `NA` BLUEs are silently dropped from GEBV estimation. |
| ID mismatch | Intersection of phenotyped and genotyped individuals is used; unmatched individuals trigger an informative message. |

The package ships with `ldx_blues` and `inst/extdata/example_blues.csv` showing
the expected format (columns: `id`, `YLD`, `RES`):

```{r blues-extdata}
blues_file <- system.file("extdata", "example_blues.csv", package = "LDxBlocks")
blues      <- read.csv(blues_file)
head(blues, 3)
```

The most common source of errors is an ID mismatch between the phenotype file
and the genotype matrix. Verify with:

```{r id-check, eval=FALSE}
data(ldx_geno)
all(blues$id %in% rownames(ldx_geno))  # should be TRUE
```

---

## 6. Step 3 — LD block detection

### 6.1 Genome-wide (recommended)

```{r run-all-chr}
blocks <- run_Big_LD_all_chr(
  ldx_geno,
  snp_info    = ldx_snp_info,
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 100L,
  n_threads   = 1L,
  verbose     = FALSE
)
nrow(blocks)
head(blocks)
```

### 6.2 Single chromosome

Use the `chr` parameter of `run_Big_LD_all_chr()` to restrict detection to
one or more specific chromosomes. This uses the same streaming memory model
and produces the same output columns as a genome-wide run.

```{r big-ld-single}
blocks_chr1 <- run_Big_LD_all_chr(
  ldx_geno,
  snp_info    = ldx_snp_info,
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 100L,
  chr         = "1",         # restrict to chromosome 1 only
  n_threads   = 1L,
  verbose     = FALSE
)
nrow(blocks_chr1)
unique(blocks_chr1$CHR)      # confirms only chr 1 processed
```

### 6.3 WGS panels: CLQmode = "Leiden" and max_bp_distance

For WGS-density panels (> 500k SNPs), the default `CLQmode = "Density"`
(Bron-Kerbosch clique enumeration) can suffer exponential blowup — a single
1500-SNP window with dense LD can stall for hours. Two parameters fix this:

```{r wgs-accel, eval=FALSE}
# Recommended for WGS panels (> 500 k SNPs)
blocks_wgs <- run_Big_LD_all_chr(
  be,
  CLQmode         = "Leiden",    # polynomial O(n log n) — guaranteed connected
  max_bp_distance = 500000L,     # skip r² for pairs > 500 kb (near-O(p))
  CLQcut          = 0.70,        # sparser LD graph
  subSegmSize     = 500L,        # smaller windows
  leng            = 50L,         # narrower boundary scan at WGS density
  checkLargest    = TRUE,        # belt-and-suspenders guard
  n_threads       = n_threads
)
```

---

## 7. Step 4 — Summarising and visualising blocks

```{r summarise}
summarise_blocks(blocks)
```

```{r plot, fig.cap="LD block structure coloured by block size"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(plot_ld_blocks(blocks, colour_by = "length_bp", mb_scale = TRUE))
}
```

---

## 8. Step 5 — Haplotype extraction

### 8.1 Unphased mode (default)

`extract_haplotypes()` accepts either a numeric dosage matrix (0/1/2/NA) or a
phased list from `read_phased_vcf()` / `phase_with_pedigree()`. In unphased
mode each individual receives one diploid allele string per block:

```{r haplotypes-unphased}
# For large real datasets, pass the backend object directly:
#   haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5L)
# This streams one chromosome at a time -- the full genome is never in RAM.
# Haplotype strings are built via build_hap_strings_cpp() (C++), replacing
# an R vapply() loop that incurred one call per individual per block.
# For a 3M-SNP panel with 17k blocks x 204 individuals: ~20-50x speedup.
#
# For the small example dataset we use the in-memory matrix path:
haps <- extract_haplotypes(
  geno     = ldx_geno,
  snp_info = ldx_snp_info,
  blocks   = blocks,
  min_snps = 5L,
  na_char  = "."
)
length(haps)
attr(haps, "block_info")
head(haps[[1]])
```

### 8.2 Phased mode

When WGS data have been phased externally (Beagle, SHAPEIT, pedigree), load
the phased VCF first and pass the resulting list to `extract_haplotypes()`.
Phased mode is detected automatically from the `hap1`/`hap2` structure:

```{r phased-mode, eval=FALSE}
# Statistical phasing via Beagle 5.x
phase_with_beagle(input_vcf = "raw.vcf.gz", out_prefix = "phased",
                  nthreads = 8L)
phased <- read_phased_vcf("phased.vcf.gz", min_maf = 0.05)

# Pedigree-based phasing
ped    <- data.frame(id = c("off1","sire1","dam1"),
                     sire = c("sire1", NA, NA),
                     dam  = c("dam1",  NA, NA))
phased <- phase_with_pedigree(ldx_geno, pedigree = ped)

# Extract haplotypes — phased strings: "011|100"
haps_p <- extract_haplotypes(phased, ldx_snp_info, blocks, min_snps = 5)
head(haps_p[[1]])   # "011|100" format
```

---

## 9. Step 6 — Haplotype diversity

`compute_haplotype_diversity()` works with both phased and unphased strings.
For phased data each individual contributes two gamete observations, doubling
the effective sample size for frequency estimates:

```{r diversity}
div <- compute_haplotype_diversity(haps)
head(div)
```

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| Richness ($k$) | unique strings | High = diverse |
| $H_e$ | Nei (1973) corrected | Expected heterozygosity |
| Shannon ($H'$) | $-\sum p_i \log p_i$ | Entropy |
| $n_{\text{eff}}$ | $1/\sum p_i^2$ | Effective number of alleles |
| $f_{\max}$ | $\max_i p_i$ | Near 1.0 = possible sweep |
| `sweep_flag` | $f_{\max} \geq 0.90$ | Logical sweep indicator |

```{r diversity-plot, fig.cap="Expected heterozygosity per LD block"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(div, ggplot2::aes(x = block_id, y = He)) +
    ggplot2::geom_col(fill = "#1D9E75") +
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle=90, size=8)) +
    ggplot2::labs(x=NULL, y="He", title="Haplotype diversity per LD block")
}
```

Write diversity results to disk:

```{r div-write, eval=FALSE}
write_haplotype_diversity(div, "haplotype_diversity.csv", append_summary = TRUE)
```

---

## 10. Step 7 — Post-GWAS QTL region definition

When GWAS results are available, map significant markers onto LD blocks to
define QTL regions and flag pleiotropic blocks:

```{r qtl-regions}
# ldx_gwas has a `trait` column (TraitA / TraitB) enabling pleiotropic detection.
# For single-trait GWAS simply omit trait_col — the function still works,
# setting n_traits = 1 and pleiotropic = FALSE for all blocks.
qtl <- define_qtl_regions(
  gwas_results = ldx_gwas,
  blocks       = blocks,
  snp_info     = ldx_snp_info,
  p_threshold  = 5e-8,
  trait_col    = "trait"      # omit this argument for single-trait GWAS
)
head(qtl[, c("block_id","CHR","n_snps_block","n_sig_markers",
             "lead_snp","traits","pleiotropic")])
subset(qtl, pleiotropic)      # blocks with hits from both TraitA and TraitB
```

---

## 11. Step 8 — Feature matrix for genomic prediction

`build_haplotype_feature_matrix()` supports two encoding schemes and both
phased and unphased input:

```{r features}
# Additive 0/1/2 (phased) or 0/2 (unphased) — recommended for GBLUP/rrBLUP
feat_add <- build_haplotype_feature_matrix(
  haplotypes = haps,
  top_n      = 5L,
  encoding   = "additive_012",
  scale_features = TRUE
)$matrix
dim(feat_add)

# Presence/absence 0/1 — for kernel methods or random forest
feat_pa <- build_haplotype_feature_matrix(
  haplotypes = haps,
  top_n      = 5L,
  encoding   = "presence_01"
)$matrix
dim(feat_pa)
```

Write the feature matrix:

```{r feat-write, eval=FALSE}
# Numeric: rows=haplotypes, cols=individuals, values=0/1/2/NA dosage
write_haplotype_numeric(feat_add, "hap_matrix_numeric.csv",
                         haplotypes = haps, snp_info = ldx_snp_info)

# Character: nucleotide sequence per individual per haplotype allele
write_haplotype_character(haplotypes = haps, snp_info = ldx_snp_info,
                           out_file = "hap_matrix_character.txt")
```

### 11.1 Haplotype prediction pipeline (Tong et al. 2025)

When pre-adjusted phenotype values (BLUEs or adjusted entry means)
are available, the full Tong et al. (2025) haplotype stacking pipeline
runs in a single call. `blues` accepts any of the four formats described
in Step 2 (named numeric vector, single-trait data frame, multi-trait
data frame, or named list). See Step 2 for column name requirements:

```{r hap-prediction, eval=FALSE}
blues <- read.csv("blues.csv")  # columns: id, YLD
pred  <- run_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info    = ldx_snp_info,
  blocks      = blocks,
  blues       = blues,
  id_col      = "id",
  blue_col    = "YLD"
)
pred$block_importance[pred$block_importance$important, ]
sort(pred$gebv, decreasing = TRUE)  # ranked GEBVs

# Integrate GWAS + variance + diversity evidence
priority <- rank_haplotype_blocks(
  diversity   = pred$diversity,
  qtl_regions = qtl,          # from define_qtl_regions()
  pred_result = pred
)
priority$ranked_blocks[priority$ranked_blocks$rank_score >= 0.9, ]
```

### 11.2 Multi-trait prediction

Pass a data frame with multiple trait columns — or a named list of named
numeric vectors — to `run_haplotype_prediction()` to analyse all traits
simultaneously with a single shared GRM. All traits are fitted via
`rrBLUP::kin.blup()` per trait using the shared GRM, making cross-trait
block importance values directly comparable. Block importance is then
aggregated across traits so rankings are not specific to any single trait:

```{r hap-prediction-mt, eval=FALSE}
blues_df <- read.csv("blues.csv")  # columns: id, YLD, DIS, PHT
res_mt <- run_haplotype_prediction(
  geno_matrix     = ldx_geno,
  snp_info        = ldx_snp_info,
  blocks          = blocks,
  blues           = blues_df,
  id_col          = "id",
  blue_cols       = c("YLD", "DIS", "PHT"),
  importance_rule = "any"   # flag block if important for >= 1 trait
)
res_mt$solver_used   # 'rrBLUP'
# Cross-trait block importance
res_mt$block_importance[res_mt$block_importance$important_any, 
  c("block_id", "var_scaled_YLD", "var_scaled_DIS",
    "var_scaled_mean", "n_traits_important")]
```

### 11.3 Building a haplotype GRM for GBLUP

```{r hap-grm}
G_hap <- tcrossprod(feat_add) / ncol(feat_add)
dim(G_hap)
round(range(diag(G_hap)), 3)
```

---

## 11. Step 8b — LD decay analysis

`compute_ld_decay()` estimates the chromosome-specific distance at which LD
drops below a critical threshold. This decay distance can then be passed to
`define_qtl_regions()` to replace fixed block boundaries with biologically
justified candidate gene windows.

```{r ld-decay, eval=FALSE}
# Memory-efficient: loads only sampled SNP columns via read_chunk()
# For the small example dataset:
decay <- compute_ld_decay(
  geno         = ldx_geno,
  snp_info     = ldx_snp_info,
  sampling     = "random",
  r2_threshold = "both",        # fixed 0.1 + parametric (95th pctile unlinked)
  fit_model    = "loess",
  n_pairs      = 3000L,
  verbose      = FALSE
)
decay$critical_r2_fixed   # 0.1 (standard GWAS threshold)
decay$critical_r2_param   # background kinship-induced LD level
decay$decay_dist          # per-chromosome decay distances

# Pass decay distances to define_qtl_regions for LD-aware windows
qtl_ld <- define_qtl_regions(
  ldx_gwas, ldx_blocks, ldx_snp_info,
  ld_decay    = decay,    # uses chromosome-specific decay distances
  p_threshold = NULL,
  trait_col   = "trait"
)
# New columns: candidate_region_start, candidate_region_end, candidate_region_size_kb
head(qtl_ld[, c("block_id", "lead_snp", "candidate_region_start",
                 "candidate_region_end", "candidate_region_size_kb")])
```

```{r ld-decay-plot, eval=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE))
  print(plot_ld_decay(decay, plot_threshold = TRUE, plot_decay_dist = TRUE))
```

---

## 12. Step 9 — Parameter auto-tuning

```{r tuning}
grid <- expand.grid(
  CLQcut = c(0.45, 0.55, 0.65), clstgap = 1e5, leng = 15,
  subSegmSize = 100, split = FALSE, checkLargest = FALSE,
  MAFcut = 0.05, CLQmode = "Density", kin_method = "chol",
  digits = -1, appendrare = FALSE, stringsAsFactors = FALSE
)

result <- tune_LD_params(
  geno_matrix    = ldx_geno,
  snp_info       = ldx_snp_info,
  gwas_df        = ldx_gwas,
  grid           = grid,
  prefer_perfect = TRUE,
  target_bp_band = c(5e4, 5e5),
  seed           = 42L
)
result$best_params
result$score_table[, c("CLQcut","n_unassigned","n_forced","n_blocks")]
```

---

## 13. Complete pipeline (one block)

```{r complete, eval=FALSE}
library(LDxBlocks)

# Step 1. Open streaming backend (VCF auto-converts to GDS cache on first call)
be     <- read_geno("mydata.vcf.gz")

# Step 2. Prepare phenotype data (BLUEs from mixed model — see Step 2 of this vignette)
blues <- read.csv("blues.csv")  # columns: id (or any name), trait1, trait2 ...

# Step 3. Block detection
# Use Leiden for WGS panels: polynomial O(n log n) and guaranteed
# connected communities (Traag et al. 2019). Louvain can produce
# disconnected communities and is not recommended.
blocks <- run_Big_LD_all_chr(
  be,
  method          = "r2",
  CLQcut          = 0.70,
  CLQmode         = "Leiden",     # polynomial; guaranteed connected communities
  max_bp_distance = 500000L,      # skip pairs > 500 kb; near-O(p)
  subSegmSize     = 500L,
  leng            = 50L,
  n_threads       = 8L
)
summarise_blocks(blocks)

# 3. Haplotypes — pass backend directly: streams one chromosome at a time,
#    full genome is NEVER loaded into RAM simultaneously.
haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)
write_haplotype_diversity(div, "diversity.csv")

feat <- build_haplotype_feature_matrix(haps,
                                        encoding = "additive_012",
                                        scale_features = TRUE)$matrix

# Numeric dosage matrix: rows=haplotypes, cols=individuals
# values: 0/1/2/NA for phased data; 0/1/NA for unphased data
write_haplotype_numeric(feat, "hap_matrix_numeric.csv",
                         haplotypes = haps, snp_info = be$snp_info)

# Nucleotide character matrix: shows actual allele sequence per individual
write_haplotype_character(haplotypes = haps, snp_info = be$snp_info,
                           out_file = "hap_matrix_character.txt")

# Decode haplotype strings to nucleotide sequences (for biological interpretation)
decoded <- decode_haplotype_strings(haps, snp_info = be$snp_info)
head(decoded[, c("block_id","hap_rank","dosage_string",
                  "nucleotide_sequence","frequency")])

close_backend(be)
```

---

## 13. Step 10 — Advanced analysis: CV, population comparison, harmonisation

The ten new analysis extension functions slot naturally after the main prediction pipeline.

### 13.1 Cross-validation

```{r cv-prediction, eval=FALSE}
cv <- cv_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info    = ldx_snp_info,
  blocks      = blocks,
  blues       = blues,
  k           = 5L,
  n_rep       = 3L,
  id_col      = "id",
  blue_col    = "YLD",
  verbose     = FALSE
)
cv$pa_mean   # mean PA and RMSE per trait across folds and replications
```

### 13.2 Population comparison

```{r pop-compare, eval=FALSE}
ids <- rownames(ldx_geno)
cmp <- compare_haplotype_populations(
  haplotypes  = haps,
  group1      = ids[1:60],
  group2      = ids[61:120],
  group1_name = "cycle1",
  group2_name = "cycle2"
)
cmp[cmp$divergent, c("block_id", "FST", "max_freq_diff", "chisq_p")]
```

### 13.3 Haplotype network for one block

```{r hap-network, eval=FALSE}
if (requireNamespace("igraph", quietly = TRUE))
  plot_haplotype_network(haps, block_id = names(haps)[1])
```

### 13.4 Diplotype inference, rare-allele collapsing, and label harmonisation

For training/validation workflows, collapse rare alleles then harmonise labels
before building feature matrices to ensure allele identity transfers correctly:

```{r harmonise, eval=FALSE}
# Build reference panel haplotypes and collapse rare alleles
haps_ref  <- extract_haplotypes(ref_geno, ldx_snp_info, blocks)
haps_ref  <- collapse_haplotypes(haps_ref, min_freq = 0.05, collapse = "nearest")

# Match validation panel alleles to the reference dictionary
haps_tgt  <- extract_haplotypes(val_geno,  ldx_snp_info, blocks)
haps_harm <- harmonize_haplotypes(haps_tgt, haps_ref, max_hamming = 3L)
attr(haps_harm, "harmonization_report")  # per-block matching quality

# Structured diplotype table
dip <- infer_block_haplotypes(haps_harm)
head(dip[, c("block_id","id","diplotype","heterozygous","phase_ambiguous")])
```

### 13.5 Candidate region export and allele effect decomposition

```{r export-effects, eval=FALSE}
# Export QTL windows as BED file
export_candidate_regions(qtl, format = "bed", chr_prefix = "chr",
                          out_file = "candidate_regions.bed")

# Decompose per-SNP effects into per-allele effects
allele_tbl <- decompose_block_effects(haps, ldx_snp_info, blocks,
                                       snp_effects = pred$snp_effects[[1]])
head(allele_tbl[order(-allele_tbl$allele_effect), ])

# Sliding-window diversity scan (independent of block boundaries)
scan <- scan_diversity_windows(ldx_geno, ldx_snp_info,
                                window_bp = 50000L, step_bp = 25000L)
```

### 13.6 Haplotype association testing

`test_block_haplotypes()` tests every LD block for association using a unified
Q+K mixed linear model: y = μ + α·x_hap + Σβ_k·PC_k + g + ε. The GRM is built
once from the haplotype feature matrix and inverted once per trait
(`rrBLUP::mixed.solve()`). Per-allele Wald tests across all blocks are then
fully vectorised in a single BLAS call. An omnibus F-test per block identifies
which blocks drive overall association.

```r
blues_vec <- setNames(ldx_blues$YLD, ldx_blues$id)

# EMMAX (pure GRM correction, n_pcs = 0 — default)
assoc <- test_block_haplotypes(
  haplotypes = haps,
  blues      = blues_vec,
  blocks     = blocks,
  n_pcs      = 0L,       # 0 = EMMAX; >0 = Q+K model
  min_freq   = 0.05,
  verbose    = FALSE
)

# Access results
cat("Total allele tests:", assoc$n_tests, "\n")
cat("Significant alleles (Bonferroni):", sum(assoc$allele_tests$significant), "\n")
cat("Significant blocks (omnibus):", sum(assoc$block_tests$significant_omnibus), "\n")

# Top blocks by omnibus p-value
head(assoc$block_tests[order(assoc$block_tests$p_omnibus),
                        c("block_id","CHR","start_bp","n_alleles_tested",
                          "F_stat","p_omnibus","var_explained")], 5)

# Significant per-allele associations with their effects
assoc$allele_tests[assoc$allele_tests$significant,
                   c("block_id","trait","allele","frequency","effect","SE","p_wald")]
```

When strong discrete population structure is present (e.g. multiple breeds or
ecotypes), use the Q+K model by setting `n_pcs > 0`. The PCs are derived from
the GRM itself, ensuring mathematical consistency between the fixed-effect
population structure correction and the kinship random effect:

```r
# Q+K model — 3 GRM-derived PCs as fixed effects + GRM kinship random effect
assoc_qk <- test_block_haplotypes(
  haplotypes = haps,
  blues      = blues_vec,
  blocks     = blocks,
  n_pcs      = 3L,
  verbose    = FALSE
)
cat("PCs used:", assoc_qk$n_pcs_used, "\n")

# Multi-trait: all traits share the same GRM — one call tests all
assoc_mt <- test_block_haplotypes(
  haplotypes = haps,
  blues      = ldx_blues,
  blocks     = blocks,
  id_col     = "id",
  blue_cols  = c("YLD", "RES"),
  n_pcs      = 3L,
  verbose    = FALSE
)
# Compare block significance across traits
library(dplyr)
assoc_mt$block_tests |>
  filter(significant_omnibus) |>
  select(block_id, CHR, trait, p_omnibus, var_explained)
```

`estimate_diplotype_effects()` decomposes phenotypic variation into additive and
dominance components at each block. The dominance ratio d/a classifies gene
action: 0 = purely additive; ±1 = complete dominance; |d/a| > 1 = overdominance
(heterozygote advantage, evidence for heterosis at this block).

```r
dip <- estimate_diplotype_effects(
  haplotypes      = haps,
  blues           = blues_vec,
  blocks          = blocks,
  min_n_diplotype = 3L,   # minimum individuals per diplotype class
  verbose         = FALSE
)

# Diplotype class means (de-regressed phenotype scale)
head(dip$diplotype_means[order(dip$diplotype_means$mean_blue, decreasing=TRUE),
                          c("block_id","diplotype","n","mean_blue","se_mean")], 8)

# Dominance decomposition: a and d for each allele pair
head(dip$dominance_table[, c("block_id","allele_A","allele_B",
                               "a","d","d_over_a","overdominance")], 8)

# Blocks showing overdominance — potential heterosis targets
od <- dip$dominance_table[!is.na(dip$dominance_table$overdominance) &
                           dip$dominance_table$overdominance, ]
cat("Overdominant allele pairs:", nrow(od), "\n")

# Statistically significant diplotype effects
dip$omnibus_tests[dip$omnibus_tests$significant,
                  c("block_id","n_diplotypes","F_stat","p_omnibus_adj")]
```

### 13.7 Breeding decision tools: haplotype stacking

Once per-allele effects are known, `score_favorable_haplotypes()` and
`summarize_parent_haplotypes()` translate genomic prediction results into
actionable selection and crossing decisions.

```r
# Step 1: Get per-allele effects from the prediction pipeline
pred <- run_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info    = ldx_snp_info,
  blocks      = blocks,
  blues       = blues_vec,
  verbose     = FALSE
)

ae <- decompose_block_effects(
  haplotypes  = haps,
  snp_info    = ldx_snp_info,
  blocks      = blocks,
  snp_effects = pred$snp_effects[[1]]   # first trait
)
cat("Allele effect table:", nrow(ae), "alleles across",
    length(unique(ae$block_id)), "blocks\n")
head(ae[order(-ae$allele_effect), c("block_id","allele","frequency","allele_effect")], 5)
```

```r
# Step 2: Score every individual's genome-wide haplotype portfolio
scores <- score_favorable_haplotypes(
  haplotypes     = haps,
  allele_effects = ae,
  min_freq       = 0.02,   # exclude private alleles
  normalize      = TRUE    # [0,1] scale within this panel
)

# Top 10 selection candidates
head(scores[, c("id","stacking_index","n_blocks_scored","rank")], 10)

# Which blocks drive the top candidate's genome-wide score?
top_candidate <- scores$id[1]
block_scores  <- scores[scores$id == top_candidate,
                         grepl("^score_", names(scores))]
top5_blocks   <- names(sort(as.numeric(block_scores), decreasing=TRUE))[1:5]
cat("Top 5 contributing blocks for", top_candidate, ":\n")
print(top5_blocks)
```

```r
# Step 3: Allele inventory for candidate parents
# — who carries what, at which blocks, in what dosage?
top10 <- scores$id[scores$rank <= 10]

inv <- summarize_parent_haplotypes(
  haplotypes     = haps,
  candidate_ids  = top10,
  allele_effects = ae,
  min_freq       = 0.02
)

# Candidates carrying rare favourable alleles (freq < 10%)
inv_rare <- inv[inv$dosage > 0 & inv$is_rare & !is.na(inv$allele_effect) &
                inv$allele_effect > 0, ]
cat("Rare favourable allele carriers:\n")
print(inv_rare[order(-inv_rare$allele_effect),
               c("id","block_id","CHR","allele","dosage","allele_freq","allele_effect")])
```

```r
# Step 4: Identify blocks where top candidates carry different alleles
# — these are the most promising crossing targets for allele stacking
library(dplyr)
inv |>
  filter(dosage > 0) |>
  group_by(block_id, CHR, start_bp) |>
  summarise(
    n_unique_alleles = n_distinct(allele),
    carriers         = paste(unique(id), collapse=","),
    max_effect       = max(allele_effect, na.rm = TRUE),
    .groups          = "drop"
  ) |>
  filter(n_unique_alleles > 1) |>            # complementary alleles present
  arrange(desc(max_effect)) |>               # rank by largest effect available
  head(10)
```

This workflow — prediction → effect decomposition → portfolio scoring →
parent inventory → crossing design — forms the complete haplotype-based
breeding decision layer implemented in LDxBlocks.

## 14. References

- Kim S-A, Cho C-S, Kim S-R, Bull SB, Yoo Y-J (2018). A new haplotype block detection method for dense genome sequencing data based on interval graph modeling and dynamic programming. *Bioinformatics* **34**(4):588-596.
  <https://doi.org/10.1093/bioinformatics/btx609>
- Difabachew YF, Frisch M, Langstroff AL, Stahl A, Wittkop B, Snowdon RJ, Koch M, Kirchhoff M, Csélényi L, Wolf M, Förster J, Weber S, Okoye UJ, Zenke-Philippi C (2023). Genomic prediction with haplotype blocks in wheat. *Frontiers in Plant Science* **14**:1168547.
  <https://doi.org/10.3389/fpls.2023.1168547>
- Weber SE, Frisch M, Snowdon RJ, Voss-Fels KP (2023). Haplotype blocks for genomic prediction: a comparative evaluation in multiple crop datasets. *Frontiers in Plant Science* **14**:1217589.
  <https://doi.org/10.3389/fpls.2023.1217589>
- Pook T, Schlather M, de los Campos G, Mayer M, Schoen CC, Simianer H (2019). HaploBlocker: Creation of subgroup-specific haplotype blocks and libraries. *Genetics* **212**(4):1045-1061.
  <https://doi.org/10.1534/genetics.119.302283>
- Tong J, Tarekegn ZT, Jambuthenne D, Alahmad S, Periyannan S, Hickey L, Dinglasan E, Hayes B (2024). Stacking beneficial haplotypes from the Vavilov wheat collection to accelerate breeding for multiple disease resistance. *Theoretical and Applied Genetics* **137**:274.
  <https://doi.org/10.1007/s00122-024-04784-w>
- Tong J et al. (2025). Haplotype stacking to improve stability of stripe rust resistance in wheat. *Theoretical and Applied Genetics* **138**:267.
  <https://doi.org/10.1007/s00122-025-05045-0>
- Mangin B, Siberchicot A, Nicolas S, Doligez A, This P, Cierco-Ayrolles C (2012). Novel measures of linkage disequilibrium that correct the bias due to population structure and relatedness. *Heredity* **108**(3):285-291.
  <https://doi.org/10.1038/hdy.2011.73>
- VanRaden PM (2008). Efficient methods to compute genomic predictions. *Journal of Dairy Science* **91**(11):4414-4423.
  <https://doi.org/10.3168/jds.2007-0980>
- Covarrubias-Pazaran G (2016). Genome-assisted prediction of quantitative traits using the R package sommer. *PLOS ONE* **11**:e0156744.
  <https://doi.org/10.1371/journal.pone.0156744>
- Calus MPL, Meuwissen THE, de Roos APW, Veerkamp RF (2008). Accuracy of genomic selection using different methods to define haplotypes. *Genetics* **178**(1):553-561.
  <https://doi.org/10.1534/genetics.107.080838>
- de Roos APW, Hayes BJ, Goddard ME (2009). Reliability of genomic predictions across multiple populations. *Genetics* **183**(4):1545-1553.
  <https://doi.org/10.1534/genetics.109.104935>
- Nei M (1973). Analysis of gene diversity in subdivided populations. *Proceedings of the National Academy of Sciences* **70**(12):3321-3323.
  <https://doi.org/10.1073/pnas.70.12.3321>
- Blondel VD, Guillaume J-L, Lambiotte R, Lefebvre E (2008). Fast unfolding of communities in large networks. *Journal of Statistical Mechanics: Theory and Experiment* **2008**:P10008.
  <https://doi.org/10.1088/1742-5468/2008/10/P10008>
- Traag VA, Waltman L, van Eck NJ (2019). From Louvain to Leiden: guaranteeing well-connected communities. *Scientific Reports* **9**:5233.
  <https://doi.org/10.1038/s41598-019-41695-z>

