---
title: "Introduction to LDxBlocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LDxBlocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
library(LDxBlocks)
data(ldx_geno,     package = "LDxBlocks")
data(ldx_snp_info, package = "LDxBlocks")
data(ldx_blocks,   package = "LDxBlocks")
data(ldx_gwas,     package = "LDxBlocks")
```

## Overview

`LDxBlocks` detects linkage disequilibrium (LD) blocks from genome-wide SNP
data and extends the pipeline into haplotype reconstruction, diversity
analysis, and genomic prediction feature engineering. Two LD metrics are
available:

- **Standard r²** -- fast, no kinship correction, suitable for 10 M+ markers
  and unstructured populations.
- **Kinship-adjusted rV²** -- corrects for relatedness-induced LD inflation,
  for livestock, inbred lines, or family-based cohorts. See the
  *LD Metrics* vignette for the statistical detail and decision table.

The C++/Armadillo computational core makes the algorithm approximately 40x
faster than the original original Big-LD implementation (no compiled code) for typical window
sizes.

This vignette covers the complete pipeline -- reading genotype data, block
detection, haplotype analysis, and parameter tuning -- using the `ldx_geno`
example dataset (120 individuals, 230 SNPs, 3 chromosomes, 9 simulated LD
blocks).

---

## Why LDxBlocks? Relationship to the original Big-LD

LDxBlocks is built on the clique-based segmentation algorithm of Kim et al.
(2018), which introduced interval graph modelling of LD bins as a principled
alternative to sliding-window approaches. The mathematical core -- CLQD bin
assignment, maximum-weight independent set block construction, and
Bron-Kerbosch clique enumeration -- is preserved exactly. LDxBlocks extends
that foundation to address three limitations of the original implementation
that become critical for modern breeding and genomics programmes.

### 1. Computational bottleneck

The original `Big_LD()` calls `cor()` inside the boundary-scan loop and once
per `CLQD()` call, with all matrix operations running through the R
interpreter. For a 50,000-SNP chromosome with the default `leng = 200` this
amounts to approximately 150,000 small R-level matrix multiplications per
chromosome. For a WGS rice panel with 3 million SNPs across 12 chromosomes,
the original implementation would take days on a workstation.

LDxBlocks replaces the critical paths with seven compiled C++ functions:

```r
# These calls are now C++/Armadillo + OpenMP -- not R loops
compute_r2_cpp(geno, digits = -1L, n_threads = 8L)   # ~40x faster
maf_filter_cpp(geno, maf_cut = 0.05)                  # ~10x faster
boundary_scan_cpp(geno, start, end, half_w, threshold) # ~20x faster
```

The outer loop of `compute_r2_cpp()` is parallelised with OpenMP. For a
1,500-SNP window with 500 individuals, the r² matrix computes in milliseconds
rather than seconds.

### 2. Memory wall

The original `Big_LD()` requires the complete genotype matrix in RAM before
detection begins. For a 10M-marker, 5,000-individual WGS dataset this is
approximately 400 GB as an R `double` matrix -- impossible on any workstation.

LDxBlocks enforces a strict never-full-genome memory model. All six supported
formats stream genotypes one chromosome window at a time:

```r
# VCF auto-converts to GDS cache; detection streams one window at a time
be     <- read_geno("wgs_panel.vcf.gz")
blocks <- run_Big_LD_all_chr(be, method = "r2", n_threads = 8L)
# Peak RAM ~ n_samples x subSegmSize x 8 bytes = 60 MB for n=5000, w=1500
```

### 3. Pipeline gap

The original Big-LD stops at the block table. For breeding programmes the
immediate next questions are: What haplotypes exist? How diverse are they?
Which blocks harbour QTLs? How do I build a multi-locus genomic prediction
model? LDxBlocks provides a complete answer:

```r
# From blocks to genomic prediction features in four lines
haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)
qtl  <- define_qtl_regions(gwas_results, blocks, be$snp_info)
feat <- build_haplotype_feature_matrix(haps, top_n = 5, encoding = "additive_012")
```

### What is preserved from the original

The following components are unchanged from Kim et al. (2018):

- `CLQD()`: bin vector assignment via maximal clique enumeration and greedy
  density/maximal priority selection
- `constructLDblock()`: maximum-weight independent set via dynamic programming
- All `CLQmode`, `clstgap`, and `split` logic
- Block table column format (`start`, `end`, `start.rsID`, `end.rsID`,
  `start.bp`, `end.bp`) -- drop-in compatible with tools that accept original
  Big-LD output

### Comparison with related tools

| Feature | Original Big-LD | gpart (Bioconductor) | PLINK --blocks | LDxBlocks |
|---|---|---|---|---|
| Algorithm | Clique/MWIS | Clique/MWIS + GPART | Gabriel et al. | Clique/MWIS |
| LD metric | r | r, D' | r² | r², rV² |
| C++ core | No | Partial | Yes | Yes (7 functions) |
| WGS streaming | No | Partial | Yes | Yes (all formats) |
| Kinship correction | No | No | No | Yes (rV²) |
| Haplotype analysis | No | No | No | Yes |
| Genomic prediction | No | No | No | Yes |
| GWAS-driven tuning | No | No | No | Yes |
| Input formats | R matrix | PLINK, VCF | PLINK | 6 formats |
| Output | Block table | Block table + gene regions | Block table | Block table + 8 downstream functions |

---

## Example data

Four datasets are shipped with the package, all generated by
`data-raw/generate_example_data.R`:

```{r example-data}
dim(ldx_geno)             # 120 individuals x 230 SNPs
head(ldx_snp_info)        # SNP, CHR, POS, REF, ALT
table(ldx_snp_info$CHR)   # chr1, chr2, chr3
nrow(ldx_gwas)            # toy GWAS markers
```

Each chromosome has three LD blocks separated by inter-block gaps and
short stretches of low-LD singleton SNPs.

---

## Step 1 — Reading genotype data

`read_geno()` auto-detects the file format from the extension and returns an
`LDxBlocks_backend` object. Six formats are supported:

| Format | Extension | `format =` |
|--------|-----------|------------|
| Numeric dosage | `.csv`, `.txt` | `"numeric"` |
| HapMap | `.hmp.txt` | `"hapmap"` |
| VCF / bgzipped VCF | `.vcf`, `.vcf.gz` | `"vcf"` |
| SeqArray GDS | `.gds` | `"gds"` |
| PLINK binary | `.bed` (+ `.bim`, `.fam`) | `"bed"` |
| R matrix | (in-memory) | `"matrix"` |

```{r read-formats, eval=FALSE}
be_vcf <- read_geno("mydata.vcf.gz")
be_csv <- read_geno("mydata.csv")
be_hmp <- read_geno("mydata.hmp.txt")
be_gds <- read_geno("mydata.gds")    # SeqArray GDS; streams SNP windows
be_bed <- read_geno("mydata.bed")    # PLINK BED; memory-mapped
```

```{r read-matrix}
be <- read_geno(ldx_geno, format = "matrix", snp_info = ldx_snp_info)
be
```

```{r backend-interface}
be$n_samples
be$n_snps
be$sample_ids[1:4]
head(be$snp_info)
chunk <- read_chunk(be, 1:30)
dim(chunk)
close_backend(be)
```

---

## Step 2 — LD block detection

### Genome-wide (recommended)

```{r run-all-chr}
blocks <- run_Big_LD_all_chr(
  ldx_geno,
  snp_info    = ldx_snp_info,
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 100L,
  n_threads   = 1L,
  verbose     = FALSE
)
nrow(blocks)
head(blocks)
```

### Single chromosome

```{r big-ld-single}
chr1_idx    <- which(ldx_snp_info$CHR == "1")
blocks_chr1 <- Big_LD(
  geno        = ldx_geno[, chr1_idx],
  SNPinfo     = ldx_snp_info[chr1_idx, c("SNP", "POS")],
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 100L,
  verbose     = FALSE
)
nrow(blocks_chr1)
```

---

## Step 3 — Summarising and visualising blocks

```{r summarise}
summarise_blocks(blocks)
```

```{r plot, fig.cap="LD block structure coloured by block size"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(plot_ld_blocks(blocks, colour_by = "length_bp", mb_scale = TRUE))
}
```

---

## Step 4 — Haplotype extraction

### Unphased mode (default)

`extract_haplotypes()` accepts either a numeric dosage matrix (0/1/2/NA) or a
phased list from `read_phased_vcf()` / `phase_with_pedigree()`. In unphased
mode each individual receives one diploid allele string per block:

```{r haplotypes-unphased}
# For large real datasets, pass the backend object directly:
#   haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5L)
# This streams one chromosome at a time — the full genome is never in RAM.
#
# For the small example dataset we use the in-memory matrix path:
haps <- extract_haplotypes(
  geno     = ldx_geno,
  snp_info = ldx_snp_info,
  blocks   = blocks,
  min_snps = 5L,
  na_char  = "."
)
length(haps)
attr(haps, "block_info")
head(haps[[1]])
```

### Phased mode

When WGS data have been phased externally (Beagle, SHAPEIT, pedigree), load
the phased VCF first and pass the resulting list to `extract_haplotypes()`.
Phased mode is detected automatically from the `hap1`/`hap2` structure:

```{r phased-mode, eval=FALSE}
# Statistical phasing via Beagle 5.x
phase_with_beagle(input_vcf = "raw.vcf.gz", out_prefix = "phased",
                  nthreads = 8L)
phased <- read_phased_vcf("phased.vcf.gz", min_maf = 0.05)

# Pedigree-based phasing
ped    <- data.frame(id = c("off1","sire1","dam1"),
                     sire = c("sire1", NA, NA),
                     dam  = c("dam1",  NA, NA))
phased <- phase_with_pedigree(ldx_geno, pedigree = ped)

# Extract haplotypes — phased strings: "011|100"
haps_p <- extract_haplotypes(phased, ldx_snp_info, blocks, min_snps = 5)
head(haps_p[[1]])   # "011|100" format
```

---

## Step 5 — Haplotype diversity

`compute_haplotype_diversity()` works with both phased and unphased strings.
For phased data each individual contributes two gamete observations, doubling
the effective sample size for frequency estimates:

```{r diversity}
div <- compute_haplotype_diversity(haps)
head(div)
```

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| Richness ($k$) | unique strings | High = diverse |
| $H_e$ | $1 - \sum p_i^2$ | Nei expected heterozygosity |
| Shannon ($H'$) | $-\sum p_i \log p_i$ | Entropy |
| $f_{\max}$ | $\max_i p_i$ | Near 1.0 = sweep |

```{r diversity-plot, fig.cap="Expected heterozygosity per LD block"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(div, ggplot2::aes(x = block_id, y = He)) +
    ggplot2::geom_col(fill = "#1D9E75") +
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle=90, size=8)) +
    ggplot2::labs(x=NULL, y="He", title="Haplotype diversity per LD block")
}
```

Write diversity results to disk:

```{r div-write, eval=FALSE}
write_haplotype_diversity(div, "haplotype_diversity.csv", append_summary = TRUE)
```

---

## Step 6 — Post-GWAS QTL region definition

When GWAS results are available, map significant markers onto LD blocks to
define QTL regions and flag pleiotropic blocks:

```{r qtl-regions}
# ldx_gwas has a `trait` column (TraitA / TraitB) enabling pleiotropic detection.
# For single-trait GWAS simply omit trait_col — the function still works,
# setting n_traits = 1 and pleiotropic = FALSE for all blocks.
qtl <- define_qtl_regions(
  gwas_results = ldx_gwas,
  blocks       = blocks,
  snp_info     = ldx_snp_info,
  p_threshold  = 5e-8,
  trait_col    = "trait"      # omit this argument for single-trait GWAS
)
head(qtl[, c("block_id","CHR","n_snps_block","n_sig_markers",
             "lead_snp","traits","pleiotropic")])
subset(qtl, pleiotropic)      # blocks with hits from both TraitA and TraitB
```

---

## Step 7 — Feature matrix for genomic prediction

`build_haplotype_feature_matrix()` supports two encoding schemes and both
phased and unphased input:

```{r features}
# Additive 0/1/2 (phased) or 0/2 (unphased) — recommended for GBLUP/rrBLUP
feat_add <- build_haplotype_feature_matrix(
  haplotypes = haps,
  top_n      = 5L,
  encoding   = "additive_012",
  scale_features = TRUE
)
dim(feat_add)

# Presence/absence 0/2 — for kernel methods or random forest
feat_pa <- build_haplotype_feature_matrix(
  haplotypes = haps,
  top_n      = 5L,
  encoding   = "presence_02"
)
dim(feat_pa)
```

Write the feature matrix in numeric or HapMap format:

```{r feat-write, eval=FALSE}
write_haplotype_numeric(feat_add, "hap_matrix_numeric.csv")
write_haplotype_hapmap(feat_add, "hap_matrix.hmp.txt")
```

### Building a haplotype GRM for GBLUP

```{r hap-grm}
G_hap <- tcrossprod(feat_add) / ncol(feat_add)
dim(G_hap)
round(range(diag(G_hap)), 3)
```

---

## Step 8 — Parameter auto-tuning

```{r tuning}
grid <- expand.grid(
  CLQcut = c(0.45, 0.55, 0.65), clstgap = 1e5, leng = 15,
  subSegmSize = 100, split = FALSE, checkLargest = FALSE,
  MAFcut = 0.05, CLQmode = "Density", kin_method = "chol",
  digits = -1, appendrare = FALSE, stringsAsFactors = FALSE
)

result <- tune_LD_params(
  geno_matrix    = ldx_geno,
  snp_info       = ldx_snp_info,
  gwas_df        = ldx_gwas,
  grid           = grid,
  prefer_perfect = TRUE,
  target_bp_band = c(5e4, 5e5),
  seed           = 42L
)
result$best_params
result$score_table[, c("CLQcut","n_unassigned","n_forced","n_blocks")]
```

---

## Complete pipeline (one block)

```{r complete, eval=FALSE}
library(LDxBlocks)

# 1. Open streaming backend (VCF auto-converts to GDS cache on first call)
be     <- read_geno("mydata.vcf.gz")

# 2. Block detection — chromosome-by-chromosome, gc() after each
blocks <- run_Big_LD_all_chr(be, method = "r2", CLQcut = 0.70, n_threads = 8L)
summarise_blocks(blocks)

# 3. Haplotypes — pass backend directly: streams one chromosome at a time,
#    full genome is NEVER loaded into RAM simultaneously.
haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)
write_haplotype_diversity(div, "diversity.csv")

feat <- build_haplotype_feature_matrix(haps, top_n = 5,
                                        encoding = "additive_012",
                                        scale_features = TRUE)
write_haplotype_numeric(feat, "hap_matrix.csv")
write_haplotype_hapmap(feat, "hap_matrix.hmp.txt")

close_backend(be)
```

---

## References

- Kim S-A et al. (2018) *Bioinformatics* **34**(4):588–596.
- VanRaden PM (2008) *J. Dairy Sci.* **91**(11):4414–4423.
- Calus MPL et al. (2008) *Genetics* **178**(1):553–561.
- de Roos APW et al. (2009) *Genetics* **183**(4):1545–1553.
- Nei M (1973) *PNAS* **70**(12):3321–3323.
- Tong J et al. (2024) *Theor Appl Genet* **137**:274.
  <https://doi.org/10.1007/s00122-024-04784-w>
