---
title: "Introduction to LDxBlocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LDxBlocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
library(LDxBlocks)
data(ldx_geno,     package = "LDxBlocks")
data(ldx_snp_info, package = "LDxBlocks")
data(ldx_blocks,   package = "LDxBlocks")
data(ldx_gwas,     package = "LDxBlocks")
```

## Overview

`LDxBlocks` detects linkage disequilibrium (LD) blocks from genome-wide SNP
data and extends the pipeline into haplotype reconstruction, diversity
analysis, and genomic prediction feature engineering. Two LD metrics are
available:

- **Standard r²** — fast, no kinship correction, suitable for 10 M+ markers
  and unstructured populations.
- **Kinship-adjusted rV²** — corrects for relatedness-induced LD inflation,
  for livestock, inbred lines, or family-based cohorts. See the
  *LD Metrics* vignette for the statistical detail and decision table.

The C++/Armadillo computational core makes the algorithm 10–50× faster than
equivalent pure-R implementations.

This vignette covers the complete pipeline — reading genotype data, block
detection, haplotype analysis, and parameter tuning — using the `ldx_geno`
example dataset (120 individuals, 200 SNPs, 3 chromosomes, 9 simulated LD
blocks).

## Example data

Four datasets are shipped with the package, all generated by
`data-raw/generate_example_data.R`:

```{r example-data}
dim(ldx_geno)             # 120 individuals x 200 SNPs
head(ldx_snp_info)        # SNP, CHR, POS, REF, ALT
table(ldx_snp_info$CHR)   # 70 on chr1, 70 on chr2, 60 on chr3
nrow(ldx_gwas)            # 20 toy GWAS markers
```

Each chromosome has three LD blocks separated by ~50 kb inter-block gaps and
short stretches of low-LD singleton SNPs.

---

## Step 1 — Reading genotype data

`read_geno()` auto-detects the file format from the extension and returns an
`LDxBlocks_backend` object. Six formats are supported:

| Format | Extension | `format =` |
|--------|-----------|------------|
| Numeric dosage | `.csv`, `.txt` | `"numeric"` |
| HapMap | `.hmp.txt` | `"hapmap"` |
| VCF / bgzipped VCF | `.vcf`, `.vcf.gz` | `"vcf"` |
| SeqArray GDS | `.gds` | `"gds"` |
| PLINK binary | `.bed` (+ `.bim`, `.fam`) | `"bed"` |
| R matrix | (in-memory) | `"matrix"` |

```{r read-formats, eval=FALSE}
# From file — format auto-detected
be_vcf <- read_geno("mydata.vcf.gz")
be_csv <- read_geno("mydata.csv")
be_hmp <- read_geno("mydata.hmp.txt")
be_gds <- read_geno("mydata.gds")    # SeqArray GDS; streams SNP windows
be_bed <- read_geno("mydata.bed")    # PLINK BED; memory-mapped
```

```{r read-matrix}
# From an in-memory R matrix (existing workflow, unchanged)
be <- read_geno(ldx_geno, format = "matrix", snp_info = ldx_snp_info)
be
```

The backend has a uniform interface regardless of source format:

```{r backend-interface}
be$n_samples        # 120
be$n_snps           # 200
be$sample_ids[1:4]
head(be$snp_info)

# Extract a genotype slice
chunk <- read_chunk(be, 1:30)
dim(chunk)          # 120 x 30

close_backend(be)   # no-op for matrix; releases handles for GDS/BED
```

---

## Step 2 — LD block detection

### Genome-wide (recommended)

```{r run-all-chr}
blocks <- run_Big_LD_all_chr(
  ldx_geno,
  snp_info    = ldx_snp_info,
  method      = "r2",      # "r2" (default) or "rV2"
  CLQcut      = 0.55,      # r² threshold for clique edges
  leng        = 15L,       # boundary-scan half-window (SNPs)
  subSegmSize = 70L,       # max SNPs per CLQD call
  n_threads   = 1L,        # increase for multi-core systems
  verbose     = FALSE
)

nrow(blocks)
head(blocks)
```

Output column definitions:

| Column | Description |
|--------|-------------|
| `start` / `end` | SNP index in the full set (including monomorphics) |
| `start.rsID` / `end.rsID` | SNP identifiers at block boundaries |
| `start.bp` / `end.bp` | Base-pair positions |
| `CHR` | Chromosome label (normalised, no `chr` prefix) |
| `length_bp` | `end.bp - start.bp + 1` |

`run_Big_LD_all_chr()` also accepts an `LDxBlocks_backend` directly, enabling
streaming access for large file-backed datasets:

```{r run-backend, eval=FALSE}
be     <- read_geno("mydata.gds")
blocks <- run_Big_LD_all_chr(be, method = "r2", CLQcut = 0.70,
                              n_threads = 8L, verbose = TRUE)
close_backend(be)
```

### Single chromosome

`Big_LD()` operates on one chromosome at a time, useful for debugging or
fine-grained control:

```{r big-ld-single}
chr1_idx    <- which(ldx_snp_info$CHR == "1")
blocks_chr1 <- Big_LD(
  geno        = ldx_geno[, chr1_idx],
  SNPinfo     = ldx_snp_info[chr1_idx, c("SNP", "POS")],
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 70L,
  verbose     = FALSE
)
nrow(blocks_chr1)
```

---

## Step 3 — Summarising and visualising blocks

```{r summarise}
summarise_blocks(blocks)
```

```{r plot, fig.cap="LD block structure coloured by block size (log10 scale)"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(plot_ld_blocks(blocks, colour_by = "length_bp", mb_scale = TRUE))
}
```

---

## Step 4 — Haplotype extraction

Within each LD block SNPs co-segregate as a unit. `extract_haplotypes()`
concatenates each individual's allele codes (0, 1, or 2) at all block SNPs
into a **phase-free diploid allele string**:

```
Individual i, block covering SNPs j1 j2 j3 j4 j5:
  haplotype = paste0(g[i, j1..j5]) = "02101"
```

Within a high-LD block these strings are nearly 1:1 with true gametic
haplotype classes (Calus et al. 2008) and are fully sufficient for diversity
metrics and prediction features. If true gametic phases are required, phase
first with SHAPEIT or Beagle and convert to 0/1/2 dosages before reading with
`read_geno()`.

```{r haplotypes}
haps <- extract_haplotypes(
  geno     = ldx_geno,
  snp_info = ldx_snp_info,
  blocks   = blocks,
  min_snps = 5L,     # skip blocks with fewer than 5 SNPs
  na_char  = "."     # placeholder for missing genotypes
)

length(haps)              # one element per qualifying block
attr(haps, "block_info")  # CHR, positions, n_snps per block
head(haps[[1]])           # haplotype strings for first block
```

---

## Step 5 — Haplotype diversity

`compute_haplotype_diversity()` computes four metrics per block:

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| Richness ($k$) | count of unique strings | High = diverse; low = selective sweep |
| $H_e$ | $\frac{n}{n-1}(1 - \sum p_i^2)$ | Nei (1973) expected heterozygosity |
| Shannon ($H'$) | $-\sum p_i \log_2 p_i$ | Entropy in bits; sensitive to evenness |
| $f_{\max}$ | $\max_i p_i$ | Near 1.0 signals a sweep or founder effect |

```{r diversity}
div <- compute_haplotype_diversity(haps)
head(div)
```

```{r diversity-plot, fig.cap="Expected heterozygosity per LD block"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(div, ggplot2::aes(x = block_id, y = He)) +
    ggplot2::geom_col(fill = "#1D9E75") +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, size = 8)
    ) +
    ggplot2::labs(x = NULL, y = "He",
                  title = "Haplotype diversity per LD block")
}
```

```{r diversity-scatter, fig.cap="Diversity landscape: dominant frequency vs He"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(div, ggplot2::aes(x = freq_dominant, y = He,
                                     colour = n_haplotypes)) +
    ggplot2::geom_point(size = 3) +
    ggplot2::scale_colour_viridis_c(option = "plasma", name = "Richness") +
    ggplot2::theme_minimal() +
    ggplot2::labs(x = "Dominant haplotype frequency", y = "He",
                  title = "Diversity landscape of LD blocks")
}
```

---

## Step 6 — Feature matrix for genomic prediction

`build_haplotype_feature_matrix()` converts haplotype strings to numeric
dosage features. For each block, the `top_n` most frequent haplotypes become
reference columns. Each individual receives 2 (homozygous match) or 0
(absent). Haplotype dosages implicitly encode multi-locus epistatic effects
without explicit interaction terms, often improving prediction accuracy by
2–8% over single-SNP models (Calus et al. 2008; de Roos et al. 2009).

```{r features}
feat <- build_haplotype_feature_matrix(
  haplotypes    = haps,
  top_n         = 5L,         # top-5 haplotypes per block as features
  scale_features = TRUE       # centre and scale each column
)

dim(feat)           # 120 x (n_blocks * 5)
feat[1:4, 1:6]
```

### Building a haplotype GRM for GBLUP

```{r hap-grm}
G_hap <- tcrossprod(feat) / ncol(feat)
dim(G_hap)                         # 120 x 120
round(range(diag(G_hap)), 3)       # diagonal values ≈ 1

# Compare to SNP-based GRM
Gc    <- scale(ldx_geno, center = TRUE, scale = FALSE)
p_snp <- colMeans(ldx_geno) / 2
G_snp <- tcrossprod(Gc) / (2 * sum(p_snp * (1 - p_snp)))
cat("Correlation between GRMs:",
    round(cor(G_hap[lower.tri(G_hap)],
              G_snp[lower.tri(G_snp)]), 3), "\n")
```

---

## Step 7 — Parameter auto-tuning

When GWAS-significant markers are available, `tune_LD_params()` selects
`CLQcut` (and optionally other parameters) that minimises, in priority order:

1. **Unassigned markers** — not falling in any block boundary.
2. **Forced assignments** — nearest-block fall-back (flagged with `*`).
3. **Number of blocks** — parsimony among equal-scoring combinations.
4. **Deviation from `target_bp_band`** — biological plausibility
   (default 50 kb – 500 kb).

When `prefer_perfect = TRUE` (default), combinations achieving zero unassigned
and zero forced are isolated first; the winner is then chosen from that subset
by criteria 3 and 4.

```{r tuning}
# Compact 3-point grid for illustration
# Use the default grid (CLQcut in {0.65, 0.70, 0.75, 0.80}) in practice
grid <- expand.grid(
  CLQcut = c(0.45, 0.55, 0.65), clstgap = 1e5, leng = 15,
  subSegmSize = 70, split = FALSE, checkLargest = FALSE,
  MAFcut = 0.05, CLQmode = "Density", kin_method = "chol",
  digits = -1, appendrare = FALSE, stringsAsFactors = FALSE
)

result <- tune_LD_params(
  geno_matrix    = ldx_geno,
  snp_info       = ldx_snp_info,
  gwas_df        = ldx_gwas,
  grid           = grid,
  prefer_perfect = TRUE,
  target_bp_band = c(5e4, 5e5),
  seed           = 42L
)
```

```{r tuning-results}
# Score for each parameter combination
result$score_table[, c("CLQcut", "n_unassigned", "n_forced",
                        "n_blocks", "median_block_bp")]

# Best parameters selected
result$best_params

# GWAS marker assignments
head(result$gwas_assigned[, c("Marker", "CHR", "POS", "LD_block")])
```

When no combination achieves zero unassigned, the most common causes are:
the marker is in a recombination hotspot with no detectable LD structure; its
MAF is below `MAFcut` and was filtered out; or the panel is sparse in that
region. Reduce `MAFcut` or set `appendrare = TRUE` in the grid to address
the second cause.

Use `result$final_blocks` as the block table for all downstream analyses:

```{r tuning-downstream}
haps_tuned <- extract_haplotypes(ldx_geno, ldx_snp_info,
                                  result$final_blocks, min_snps = 5L)
feat_tuned <- build_haplotype_feature_matrix(haps_tuned, top_n = 5,
                                              scale_features = TRUE)
cat("Blocks:", nrow(result$final_blocks),
    "| Features:", ncol(feat_tuned), "\n")
```

For parallel grid search across combinations:

```{r parallel-tune, eval=FALSE}
library(future)
plan(multisession, workers = 4L)
result_par <- tune_LD_params(ldx_geno, ldx_snp_info, ldx_gwas,
                              parallel = TRUE, seed = 42L)
plan(sequential)
```

---

## Complete pipeline (one block)

```{r complete, eval=FALSE}
library(LDxBlocks)

# 1. Read
be <- read_geno("mydata.vcf.gz")

# 2. Detect blocks
blocks <- run_Big_LD_all_chr(be, method = "r2", CLQcut = 0.70,
                              n_threads = 8L)

# 3. Summarise
summarise_blocks(blocks)
plot_ld_blocks(blocks)

# 4. Haplotypes + diversity
geno_all <- read_chunk(be, 1:be$n_snps)
haps     <- extract_haplotypes(geno_all, be$snp_info, blocks, min_snps = 5)
div      <- compute_haplotype_diversity(haps)

# 5. Prediction features
feat <- build_haplotype_feature_matrix(haps, top_n = 5, scale_features = TRUE)

# 6. Release
close_backend(be)
```

---

## References

- Kim S-A et al. (2018) *Bioinformatics* **34**(4):588–596.
  <https://doi.org/10.1093/bioinformatics/btx609>
- VanRaden PM (2008) *J. Dairy Sci.* **91**(11):4414–4423.
  <https://doi.org/10.3168/jds.2007-0980>
- Calus MPL et al. (2008) *Genetics* **178**(1):553–561.
  <https://doi.org/10.1534/genetics.107.080838>
- de Roos APW et al. (2009) *Genetics* **183**(4):1545–1553.
  <https://doi.org/10.1534/genetics.109.104935>
- Nei M (1973) *PNAS* **70**(12):3321–3323.
  <https://doi.org/10.1073/pnas.70.12.3321>
