---
title: "LD Metrics: Standard r² and Kinship-Adjusted rV²"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LD Metrics: Standard r² and Kinship-Adjusted rV²}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message   = FALSE,
  warning   = FALSE
)
library(LDxBlocks)
data(ldx_geno,     package = "LDxBlocks")
data(ldx_snp_info, package = "LDxBlocks")
```

## 1. The kinship problem in LD estimation

The standard squared Pearson correlation between two SNP dosage vectors
$g_j$ and $g_k$ is:

$$r^2_{jk} = \left[\frac{\mathrm{Cov}(g_j, g_k)}
  {\sqrt{\mathrm{Var}(g_j)\,\mathrm{Var}(g_k)}}\right]^2$$

This estimator assumes the $n$ individuals are drawn independently from the
same population. In structured or related populations this fails: two
individuals sharing a recent common ancestor carry correlated alleles at
*all* loci, not just loci in true LD. The sample covariance picks up both
genuine LD *and* kinship-induced allele sharing, inflating $r^2$ and causing
blocks to be drawn too broadly.

## 2. The kinship correction: rV²

Mangin et al. (2012) proposed the **kinship-adjusted squared correlation**:

$$rV^2_{jk} = \left[\mathrm{Cor}(V^{-1/2} g_j,\; V^{-1/2} g_k)\right]^2$$

where $V = ZZ^\top / (2\sum_j p_j q_j)$ is the VanRaden (2008) GRM and
$V^{-1/2}$ is its inverse square root. Left-multiplying by $V^{-1/2}$ removes
kinship-induced correlation so the resulting correlation measures only
recombination-based LD.

## 3. Decision table

```{r decision-table, echo=FALSE}
dt <- data.frame(
  Criterion = c("Population type", "Computational cost",
                "RAM requirement", "Marker scale",
                "Block accuracy in related pops",
                "External dependencies"),
  `r² (method = 'r2')` = c(
    "Random mating, unrelated, weakly structured",
    "O(np) prep + O(p²) per window via C++",
    "Proportional to one window",
    "Scales to 10 M+ markers",
    "Slightly inflated — blocks too broad",
    "None (always installed)"
  ),
  `rV² (method = 'rV2')` = c(
    "Livestock, inbred lines, family-based cohorts",
    "O(n²p) GRM + O(n³) Cholesky + O(np) whitening",
    "n×n GRM + n×n whitening factor per chromosome",
    "Practical to ~200 k markers per chromosome",
    "Correct for structured populations",
    "AGHmatrix, ASRgenomics (optional Suggests)"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(dt, caption = "r² versus rV²: decision table")
```

## 4. Computing rV² in LDxBlocks

```{r prepare-demo}
prep_r2 <- prepare_geno(ldx_geno[, 1:30], method = "r2")
str(prep_r2)
```

### 4.1 The VanRaden GRM

$$V = \frac{ZZ^\top}{2\sum_j p_j q_j}$$

where $z_{ij} = g_{ij} - 2p_j$ is the frequency-centred genotype.

### 4.2 The whitening factor

`get_V_inv_sqrt()` computes $A = V^{-1/2}$ such that $A V A^\top = I$.

- **Cholesky** (`kin_method = "chol"`, default): $A = R^{-1}$ where $V = R^\top R$.
- **Eigendecomposition** (`kin_method = "eigen"`): $A = Q \Lambda^{-1/2} Q^\top$.

```{r whitening-demo}
set.seed(1)
G_small <- ldx_geno[, 1:30]
Gc      <- scale(G_small, center = TRUE, scale = FALSE)
V_demo  <- tcrossprod(Gc) / 30 + diag(0.05, nrow(Gc))

A_chol <- get_V_inv_sqrt(V_demo, method = "chol")
A_eig  <- get_V_inv_sqrt(V_demo, method = "eigen")

max(abs(A_chol %*% V_demo %*% t(A_chol) - diag(120)))
max(abs(A_eig  %*% V_demo %*% t(A_eig)  - diag(120)))
```

## 5. Comparing r² and rV² on example data

```{r compare-matrices}
idx_25  <- 1:25
r2_mat  <- compute_r2(ldx_geno[, idx_25])

Gc_25   <- scale(ldx_geno[, idx_25], center = TRUE, scale = FALSE)
V_toy   <- tcrossprod(Gc_25) / 25 + diag(0.1, 120)
A_toy   <- get_V_inv_sqrt(V_toy)
X_whit  <- A_toy %*% Gc_25
rv2_mat <- compute_rV2(X_whit)

cat("Mean r² :", round(mean(r2_mat[upper.tri(r2_mat)]),  4), "\n")
cat("Mean rV²:", round(mean(rv2_mat[upper.tri(rv2_mat)]), 4), "\n")
```

```{r compare-image, fig.cap="r² (left) vs rV² (right) for the first 25 SNPs on chr1", fig.width=9}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
image(r2_mat,  main = "Standard r²",
      col = hcl.colors(20, "YlOrRd", rev = TRUE),
      xlab = "SNP", ylab = "SNP")
image(rv2_mat, main = "Kinship-adjusted rV²",
      col = hcl.colors(20, "YlOrRd", rev = TRUE),
      xlab = "SNP", ylab = "SNP")
par(mfrow = c(1, 1))
```

## 6. When r² gives wrong blocks

In a livestock panel where sires appear repeatedly through progeny, the
kinship-induced correlation between half-sibs inflates $r^2$ for all SNP
pairs in the same family cluster. A pair of SNPs on different chromosomes
can show $r^2 > 0.5$ purely because both correlate with family membership.
After left-multiplication by $V^{-1/2}$, only genuine gametic LD contributes
to $rV^2$. In practice, rV² blocks are 10–30% smaller and more precisely
delimited in related populations.

## 7. Switching between metrics

```{r switch, eval=FALSE}
blocks_r2  <- run_Big_LD_all_chr(be, method = "r2",  CLQcut = 0.70)
blocks_rv2 <- run_Big_LD_all_chr(be, method = "rV2", CLQcut = 0.70,
                                  kin_method = "chol")
```

## 8. LD decay and the parametric threshold

A high parametric threshold (> 0.05) from `compute_ld_decay()` is direct
evidence that `method = "rV2"` should be used — the same kinship inflation
demonstrated above is visible in unlinked-marker r² values.

```{r ld-decay-demo, eval=FALSE}
decay <- compute_ld_decay(ldx_geno, ldx_snp_info,
                           r2_threshold = "both", n_pairs = 2000L, verbose = FALSE)
cat("Parametric threshold:", round(decay$critical_r2_param, 4), "\n")
```

## 9. References

- Kim S-A et al. (2018). *Bioinformatics* **34**(4):588-596.
  <https://doi.org/10.1093/bioinformatics/btx609>
- Mangin B et al. (2012). *Heredity* **108**(3):285-291.
  <https://doi.org/10.1038/hdy.2011.73>
- VanRaden PM (2008). *Journal of Dairy Science* **91**(11):4414-4423.
  <https://doi.org/10.3168/jds.2007-0980>
