---
title: "Introduction to LDxBlocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LDxBlocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
library(LDxBlocks)
data(ldx_geno,     package = "LDxBlocks")
data(ldx_snp_info, package = "LDxBlocks")
data(ldx_blocks,   package = "LDxBlocks")
data(ldx_gwas,     package = "LDxBlocks")
```

## 1. Overview

`LDxBlocks` detects linkage disequilibrium (LD) blocks from genome-wide SNP
data and extends the pipeline into haplotype reconstruction, diversity
analysis, and genomic prediction feature engineering. Two LD metrics are
available:

- **Standard r²** -- fast, no kinship correction, suitable for 10 M+ markers
  and unstructured populations.
- **Kinship-adjusted rV²** -- corrects for relatedness-induced LD inflation,
  for livestock, inbred lines, or family-based cohorts. See the
  *LD Metrics* vignette for the statistical detail and decision table.

The C++/Armadillo computational core makes the algorithm approximately 40x
faster than the original Big-LD implementation for typical window sizes.

This vignette covers the complete pipeline -- reading genotype data, block
detection, haplotype analysis, and parameter tuning -- using the `ldx_geno`
example dataset (120 individuals, 230 SNPs, 3 chromosomes, 9 simulated LD
blocks).

---

## 2. Why LDxBlocks? Relationship to the original Big-LD

LDxBlocks is built on the clique-based segmentation algorithm of Kim et al.
(2018). The mathematical core -- CLQD bin assignment, maximum-weight
independent set block construction, and Bron-Kerbosch clique enumeration --
is preserved exactly. LDxBlocks extends that foundation to address three
limitations of the original implementation.

### 2.1 Computational bottleneck

The original `Big_LD()` calls `cor()` inside the boundary-scan loop and once
per `CLQD()` call, with all matrix operations running through the R
interpreter. For a WGS rice panel with 3 million SNPs across 12 chromosomes,
the original implementation would take days on a workstation.

LDxBlocks replaces the critical paths with eight compiled C++ functions:

```r
compute_r2_cpp(geno, digits = -1L, n_threads = 8L)   # ~40x faster
maf_filter_cpp(geno, maf_cut = 0.05)                  # ~10x faster
boundary_scan_cpp(geno, start, end, half_w, threshold) # ~20x faster
build_hap_strings_cpp(blk_int, na_char)               # ~20-50x faster
```

### 2.2 Memory wall

LDxBlocks enforces a strict never-full-genome memory model. All six supported
formats stream genotypes one chromosome window at a time:

```r
be     <- read_geno("wgs_panel.vcf.gz")
blocks <- run_Big_LD_all_chr(be, method = "r2", n_threads = 8L)
# Peak RAM ~ n_samples x subSegmSize x 8 bytes = 60 MB for n=5000, w=1500
```

### 2.3 Pipeline gap

The original Big-LD stops at the block table. LDxBlocks provides a complete
downstream pipeline:

```r
haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)
qtl  <- define_qtl_regions(gwas_results, blocks, be$snp_info)
feat <- build_haplotype_feature_matrix(haps, top_n = 5, encoding = "additive_012")$matrix
```

### 2.4 What is preserved from the original

- `CLQD()`: bin vector assignment via maximal clique enumeration
- `constructLDblock()`: maximum-weight independent set via dynamic programming
- All `CLQmode`, `clstgap`, and `split` logic
- Block table column format -- drop-in compatible with tools that accept
  original Big-LD output

---

## 3. Example data

```{r example-data}
dim(ldx_geno)             # 120 individuals x 230 SNPs
head(ldx_snp_info)        # SNP, CHR, POS, REF, ALT
table(ldx_snp_info$CHR)   # chr1, chr2, chr3
nrow(ldx_gwas)            # toy GWAS markers
```

---

## 4. Step 1 — Reading genotype data

```{r read-formats, eval=FALSE}
be_vcf <- read_geno("mydata.vcf.gz")
be_csv <- read_geno("mydata.csv")
be_hmp <- read_geno("mydata.hmp.txt")
be_gds <- read_geno("mydata.gds")
be_bed <- read_geno("mydata.bed")
```

```{r read-matrix}
be <- read_geno(ldx_geno, format = "matrix", snp_info = ldx_snp_info)
be
```

```{r backend-interface}
be$n_samples
be$n_snps
be$sample_ids[1:4]
head(be$snp_info)
chunk <- read_chunk(be, 1:30)
dim(chunk)
close_backend(be)
```

---

## 5. Step 2 — Phenotype input format

The `blues` argument accepts pre-adjusted phenotype means. Four formats are
accepted:

```{r blues-vector, eval=FALSE}
# Format 1: Named numeric vector
blues <- c(ind001 = 4.21, ind002 = 3.87, ind003 = 5.14)

# Format 2: Data frame, single trait
blues <- read.csv("blues.csv")
res   <- run_haplotype_prediction(geno, snp_info, blocks,
                                   blues    = blues,
                                   id_col   = "Genotype",
                                   blue_col = "YLD_BLUE")

# Format 3: Data frame, multiple traits
res_mt <- run_haplotype_prediction(geno, snp_info, blocks,
                                    blues     = blues_mt,
                                    id_col    = "id",
                                    blue_cols = c("YLD", "DIS", "PHT"))

# Format 4: Named list (different individuals per trait)
blues <- list(
  YLD = c(ind001 = 4.21, ind002 = 3.87),
  DIS = c(ind001 = 0.32, ind003 = 0.28)
)
```

```{r blues-extdata}
blues_file <- system.file("extdata", "example_blues.csv", package = "LDxBlocks")
blues      <- read.csv(blues_file)
head(blues, 3)
```

---

## 6. Step 3 — LD block detection

### 6.1 Genome-wide (recommended)

```{r run-all-chr}
blocks <- run_Big_LD_all_chr(
  ldx_geno,
  snp_info    = ldx_snp_info,
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 100L,
  n_threads   = 1L,
  verbose     = FALSE
)
nrow(blocks)
head(blocks)
```

### 6.2 Single chromosome

```{r big-ld-single}
blocks_chr1 <- run_Big_LD_all_chr(
  ldx_geno,
  snp_info    = ldx_snp_info,
  method      = "r2",
  CLQcut      = 0.55,
  leng        = 15L,
  subSegmSize = 100L,
  chr         = "1",
  n_threads   = 1L,
  verbose     = FALSE
)
nrow(blocks_chr1)
unique(blocks_chr1$CHR)
```

### 6.3 WGS panels: CLQmode = "Leiden" and max_bp_distance

```{r wgs-accel, eval=FALSE}
blocks_wgs <- run_Big_LD_all_chr(
  be,
  CLQmode         = "Leiden",
  max_bp_distance = 500000L,
  CLQcut          = 0.70,
  subSegmSize     = 500L,
  leng            = 50L,
  checkLargest    = TRUE,
  n_threads       = n_threads
)
```

---

## 7. Step 4 — Summarising and visualising blocks

```{r summarise}
summarise_blocks(blocks)
```

```{r plot, fig.cap="LD block structure coloured by block size"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(plot_ld_blocks(blocks, colour_by = "length_bp", mb_scale = TRUE))
}
```

---

## 8. Step 5 — Haplotype extraction

```{r haplotypes-unphased}
haps <- extract_haplotypes(
  geno     = ldx_geno,
  snp_info = ldx_snp_info,
  blocks   = blocks,
  min_snps = 5L,
  na_char  = "."
)
length(haps)
attr(haps, "block_info")
head(haps[[1]])
```

---

## 9. Step 6 — Haplotype diversity

```{r diversity}
div <- compute_haplotype_diversity(haps)
head(div)
```

```{r diversity-plot, fig.cap="Expected heterozygosity per LD block"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(div, ggplot2::aes(x = block_id, y = He)) +
    ggplot2::geom_col(fill = "#1D9E75") +
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle=90, size=8)) +
    ggplot2::labs(x=NULL, y="He", title="Haplotype diversity per LD block")
}
```

---

## 10. Step 7 — Post-GWAS QTL region definition

```{r qtl-regions}
qtl <- define_qtl_regions(
  gwas_results = ldx_gwas,
  blocks       = blocks,
  snp_info     = ldx_snp_info,
  p_threshold  = 5e-8,
  trait_col    = "trait"
)
head(qtl[, c("block_id","CHR","n_snps_block","n_sig_markers",
             "lead_snp","traits","pleiotropic")])
subset(qtl, pleiotropic)
```

---

## 11. Step 8 — Feature matrix for genomic prediction

```{r features}
feat_add <- build_haplotype_feature_matrix(
  haplotypes = haps,
  top_n      = 5L,
  encoding   = "additive_012",
  scale_features = TRUE
)$matrix
dim(feat_add)
```

```{r hap-grm}
G_hap <- tcrossprod(feat_add) / ncol(feat_add)
dim(G_hap)
round(range(diag(G_hap)), 3)
```

---

## 12. Step 9 — Parameter auto-tuning

```{r tuning}
grid <- expand.grid(
  CLQcut = c(0.45, 0.55, 0.65), clstgap = 1e5, leng = 15,
  subSegmSize = 100, split = FALSE, checkLargest = FALSE,
  MAFcut = 0.05, CLQmode = "Density", kin_method = "chol",
  digits = -1, appendrare = FALSE, stringsAsFactors = FALSE
)

result <- tune_LD_params(
  geno_matrix    = ldx_geno,
  snp_info       = ldx_snp_info,
  gwas_df        = ldx_gwas,
  grid           = grid,
  prefer_perfect = TRUE,
  target_bp_band = c(5e4, 5e5),
  seed           = 42L
)
result$best_params
result$score_table[, c("CLQcut","n_unassigned","n_forced","n_blocks")]
```

---

## 13. Advanced analyses

### 13.1 Cross-validation

```{r cv-prediction, eval=FALSE}
cv <- cv_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info    = ldx_snp_info,
  blocks      = blocks,
  blues       = blues,
  k           = 5L,
  n_rep       = 3L,
  id_col      = "id",
  blue_col    = "YLD",
  verbose     = FALSE
)
cv$pa_mean
```

### 13.2 Population comparison

```{r pop-compare, eval=FALSE}
ids <- rownames(ldx_geno)
cmp <- compare_haplotype_populations(
  haplotypes  = haps,
  group1      = ids[1:60],
  group2      = ids[61:120],
  group1_name = "cycle1",
  group2_name = "cycle2"
)
cmp[cmp$divergent, c("block_id", "FST", "max_freq_diff", "chisq_p")]
```

### 13.3 Diplotype inference, rare-allele collapsing, and label harmonisation

```{r harmonise, eval=FALSE}
haps_ref  <- extract_haplotypes(ref_geno, ldx_snp_info, blocks)
haps_ref  <- collapse_haplotypes(haps_ref, min_freq = 0.05, collapse = "nearest")
haps_tgt  <- extract_haplotypes(val_geno, ldx_snp_info, blocks)
haps_harm <- harmonize_haplotypes(haps_tgt, haps_ref, max_hamming = 3L)
attr(haps_harm, "harmonization_report")
dip <- infer_block_haplotypes(haps_harm)
head(dip[, c("block_id","id","diplotype","heterozygous","phase_ambiguous")])
```

### 13.4 Candidate region export and effect decomposition

```{r export-effects, eval=FALSE}
export_candidate_regions(qtl, format = "bed", chr_prefix = "chr",
                          out_file = "candidate_regions.bed")

allele_tbl <- decompose_block_effects(haps, ldx_snp_info, blocks,
                                       snp_effects = pred$snp_effects[[1]])
head(allele_tbl[order(-allele_tbl$allele_effect), ])

scan <- scan_diversity_windows(ldx_geno, ldx_snp_info,
                                window_bp = 50000L, step_bp = 25000L)
```

### 13.5 Breeding decision tools: haplotype stacking

```{r stacking, eval=FALSE}
pred <- run_haplotype_prediction(
  geno_matrix = ldx_geno, snp_info = ldx_snp_info, blocks = blocks,
  blues = blues_vec, verbose = FALSE
)
ae <- decompose_block_effects(
  haplotypes = haps, snp_info = ldx_snp_info,
  blocks = blocks, snp_effects = pred$snp_effects[[1]]
)
scores <- score_favorable_haplotypes(haps, allele_effects = ae, normalize = TRUE)
head(scores[, c("id","stacking_index","n_blocks_scored","rank")], 10)
top10  <- scores$id[scores$rank <= 10]
inv    <- summarize_parent_haplotypes(haps, candidate_ids = top10,
                                       allele_effects = ae, min_freq = 0.02)
inv[inv$dosage > 0 & inv$is_rare & !is.na(inv$allele_effect) & inv$allele_effect > 0,
    c("id","block_id","allele","dosage","allele_freq","allele_effect")]
```

### 13.6 Haplotype association testing with simpleM correction

`test_block_haplotypes()` uses a unified Q+K mixed linear model. The
**simpleM** procedure (Gao et al. 2008, 2010, 2011) estimates the effective
number of independent tests (Meff) from the eigenspectrum of the haplotype
allele dosage correlation matrix, providing family-wise error control that is
less conservative than raw Bonferroni.

```r
blues_vec <- setNames(ldx_blues$YLD, ldx_blues$id)

# EMMAX with FDR — recommended for discovery
assoc_fdr <- test_block_haplotypes(
  haplotypes = haps,
  blues      = blues_vec,
  blocks     = blocks,
  n_pcs      = 0L,
  sig_metric = "p_fdr",
  verbose    = FALSE
)
cat("Significant (FDR <= 0.05):", sum(assoc_fdr$allele_tests$significant), "\n")

# Q+K with simpleM Šidák — recommended for family-wise error control
assoc_sm <- test_block_haplotypes(
  haplotypes       = haps,
  blues            = blues_vec,
  blocks           = blocks,
  n_pcs            = 3L,                 # Q+K model (used when optimize_pcs = FALSE)
  sig_metric       = "p_simplem_sidak",  # simpleM Šidák correction
  meff_scope       = "chromosome",       # recommended: separate Meff per chr
  meff_percent_cut = 0.995,              # 99.5% variance threshold (Gao default)
  verbose          = FALSE
)

# Per-chromosome Meff
assoc_sm$meff$trait$allele$chromosome

# All four p-value columns always present regardless of sig_metric
# p_wald | p_fdr | p_simplem | p_simplem_sidak | Meff | alpha_simplem | ...
head(assoc_sm$allele_tests[order(assoc_sm$allele_tests$p_wald),
                            c("block_id","allele","effect","SE",
                              "p_wald","p_simplem","p_simplem_sidak","Meff")], 5)

# Automatic PC selection — let the data choose n_pcs via BIC/lambda hybrid
# Recommended when lambda_GC is far from 1.0 with a fixed n_pcs.
assoc_opt <- test_block_haplotypes(
  haplotypes       = haps,
  blues            = blues_vec,
  blocks           = blocks,
  optimize_pcs     = TRUE,              # triggers BIC/lambda model selection
  optimize_pcs_max = 10L,              # test k = 0..10
  optimize_method  = "bic_lambda",     # recommended: |lambda-1| + 0.01*BIC
  sig_metric       = "p_simplem_sidak",
  meff_scope       = "chromosome",
  plot             = TRUE,             # saves PDF plots including pca_grm.pdf
  out_dir          = "results",        # and grm_scree.pdf
  verbose          = TRUE
)

# Inspect selection table
print(assoc_opt$pc_model_selection)
#  n_pcs     BIC lambda_gc  score selected
#      0  1234.1     1.021 0.0210    FALSE
#      1  1231.5     1.015 0.0150    FALSE
#      2  1230.8     1.003 0.0033    FALSE
#      3  1232.1     0.999 0.0011     TRUE   <-- selected

# Multi-trait: all traits share the same GRM
assoc_mt <- test_block_haplotypes(
  haplotypes = haps,
  blues      = ldx_blues,
  blocks     = blocks,
  id_col     = "id",
  blue_cols  = c("YLD", "RES"),
  n_pcs      = 3L,
  sig_metric = "p_simplem_sidak",
  verbose    = FALSE
)
```

`estimate_diplotype_effects()` now supports the same correction set as
`test_block_haplotypes()`. All five p-value columns are always present in
`$omnibus_tests`; `sig_metric` controls which drives the `significant` flag:

```r
dip <- estimate_diplotype_effects(
  haplotypes       = haps,
  blues            = blues_vec,
  blocks           = blocks,
  min_n_diplotype  = 3L,
  sig_metric       = "p_omnibus_simplem_sidak",  # recommended
  meff_percent_cut = 0.995,
  verbose          = FALSE
)

# All five columns always present:
# p_omnibus_adj | p_omnibus_fdr | p_omnibus_simplem | p_omnibus_simplem_sidak | Meff
head(dip$omnibus_tests[, c("block_id","trait","F_stat","p_omnibus",
                            "p_omnibus_fdr","p_omnibus_simplem_sidak",
                            "Meff","significant")], 5)

# Blocks showing overdominance (heterosis candidates)
dip$dominance_table[dip$dominance_table$overdominance,
                    c("block_id","allele_A","allele_B","a","d","d_over_a")]
```

### 13.7 Breeding decision summary

See section 13.5 above and the complete example in `?score_favorable_haplotypes`.

### 13.8 Cross-population GWAS validation

`compare_block_effects()` validates whether haplotype effects from one
population replicate in another. It computes IVW meta-analytic effects,
Cochran Q heterogeneity (tests whether effect sizes differ between
populations), I² inconsistency, and direction agreement per block.

**Key distinction:** `boundary_overlap_ratio` is an **output column**
automatically computed from the block tables you supply — you cannot set it.
`boundary_overlap_warn` is an **input parameter** (default `0.80`) that
controls when `boundary_warning = TRUE` is raised in the output.

```r
# Best practice: use Pop A's block boundaries for Pop B to maximise
# shared alleles and ensure haplotype strings are directly comparable.
haps_B_harm <- harmonize_haplotypes(
  extract_haplotypes(geno_B, snp_info, blocks_A),  # same block coordinates
  reference = haps_A
)

assoc_A <- test_block_haplotypes(haps_A, blues = blues_A, blocks = blocks_A,
                                  sig_metric = "p_simplem_sidak")
assoc_B <- test_block_haplotypes(haps_B_harm, blues = blues_B, blocks = blocks_A,
                                  sig_metric = "p_simplem_sidak")

# block_match = "id" (default): match by block_id string
# block_match = "position": match by genomic IoU -- use when blocks differ
conc <- compare_block_effects(
  assoc_A, assoc_B,
  pop1_name             = "PopA",
  pop2_name             = "PopB",
  blocks_pop1           = blocks_A,
  blocks_pop2           = blocks_B,   # Pop B own block table (may differ)
  block_match           = "position", # recommended when blocks not identical
  overlap_min           = 0.50,
  direction_threshold   = 0.75,
  boundary_overlap_warn = 0.80
)

# $concordance — one row per block: IVW effect, Q_stat, I2, replicated, ...
conc$concordance[conc$concordance$replicated,
                 c("block_id","n_shared_alleles","direction_agreement",
                   "meta_p","Q_p","I2","replicated")]

# $shared_alleles — per-allele IVW detail
head(conc$shared_alleles[, c("block_id","allele","effect_pop1","effect_pop2",
                               "direction_agree","ivw_effect","ivw_SE")])

print(conc)   # summary: n blocks compared, n replicated, median I²
```

**Interpreting the output:**

| Statistic | Strong replication | Concern |
|-----------|-------------------|---------|
| `direction_agreement` | ≥ 0.75 | < 0.5 (effects inconsistent) |
| `Q_p` | > 0.05 (no heterogeneity) | < 0.05 (effect sizes differ) |
| `I2` | < 25% | > 50% (substantial heterogeneity) |
| `boundary_warning` | FALSE | TRUE (LD structure may differ) |
| `match_type` | `"exact"` or `"position"` | `"pop1_only"` (no overlapping Pop2 block) |

---


### 13.9 Cross-population validation from external GWAS

When GWAS was run outside LDxBlocks, use `compare_gwas_effects()`. It accepts
either the output of `define_qtl_regions()` (recommended, most auditable) or
raw GWAS data frames plus block tables (convenience path).

```r
# Path 1: pre-mapped (recommended)
# define_qtl_regions() maps each marker to its LD block and identifies
# the lead SNP (smallest p-value) as the block representative.
qtl_A <- define_qtl_regions(gwas_A, blocks, snp_info, p_threshold = 5e-8)
qtl_B <- define_qtl_regions(gwas_B, blocks, snp_info, p_threshold = 5e-8)

conc_gwas <- compare_gwas_effects(
  qtl_pop1    = qtl_A,
  qtl_pop2    = qtl_B,
  blocks_pop1 = blocks_A,
  blocks_pop2 = blocks_B,
  block_match = "position",
  overlap_min = 0.50,
  pop1_name   = "PopA",
  pop2_name   = "PopB"
)

# Path 2: raw GWAS + blocks (calls define_qtl_regions internally)
# When SE is absent, it is derived from BETA and P via z-score.
# Column names are flexible: use beta_col, se_col, p_col arguments.
conc_raw <- compare_gwas_effects(
  gwas_pop1     = gwas_A,
  gwas_pop2     = gwas_B,
  blocks_pop1   = blocks,
  blocks_pop2   = blocks,
  snp_info_pop1 = snp_info,   # snp_info_pop2 reuses this when NULL (shared panel)
  pop1_name     = "PopA",
  pop2_name     = "PopB",
  p_threshold   = 5e-8,
  beta_col      = "BETA",     # change if named differently in your GWAS output
  se_col        = "SE",       # NULL or absent: derived from z-score automatically
  p_col         = "P"
)

# Same output class as compare_block_effects()
# Extra columns: lead_snp_pop1/pop2, lead_p_pop1/pop2,
#                se_derived_pop1/pop2, both_pleiotropic
conc_gwas$concordance[conc_gwas$concordance$replicated, ]
print(conc_gwas)
```

Note that because external GWAS provides one lead SNP per block (not multiple
haplotype alleles), `effect_correlation` and Cochran Q are always `NA`.
Replication is judged by direction agreement and `meta_p ≤ 0.05`.

## 14. References

- Kim S-A et al. (2018). *Bioinformatics* **34**(4):588-596.
  <https://doi.org/10.1093/bioinformatics/btx609>
- Gao X, Starmer J, Martin ER (2008). A multiple testing correction method for
  genetic association studies using correlated SNPs. *Genetic Epidemiology*
  **32**:361-369. <https://doi.org/10.1002/gepi.20310>
- Gao X et al. (2010). Avoiding the high Bonferroni penalty in GWAS.
  *Genetic Epidemiology* **34**:100-105. <https://doi.org/10.1002/gepi.20430>
- Gao X (2011). Multiple testing corrections for imputed SNPs.
  *Genetic Epidemiology* **35**:154-158. <https://doi.org/10.1002/gepi.20563>
- Borenstein M et al. (2009). *Introduction to Meta-Analysis*. Wiley.
- Higgins JPT, Thompson SG (2002). Quantifying heterogeneity in a meta-analysis.
  *Statistics in Medicine* **21**:1539-1558. <https://doi.org/10.1002/sim.1186>
- Tong J et al. (2025). *Theor Appl Genet* **138**:267.
  <https://doi.org/10.1007/s00122-025-05045-0>
- Tong J et al. (2024). *Theor Appl Genet* **137**:274.
  <https://doi.org/10.1007/s00122-024-04784-w>
- Mangin B et al. (2012). *Heredity* **108**(3):285-291.
  <https://doi.org/10.1038/hdy.2011.73>
- VanRaden PM (2008). *Journal of Dairy Science* **91**(11):4414-4423.
  <https://doi.org/10.3168/jds.2007-0980>
- Nei M (1973). *PNAS* **70**(12):3321-3323.
  <https://doi.org/10.1073/pnas.70.12.3321>
- Traag VA et al. (2019). *Scientific Reports* **9**:5233.
  <https://doi.org/10.1038/s41598-019-41695-z>
