---
title: "LDxBlocks Workflow: From Genotype File to Haplotype Features"
author: "LDxBlocks Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{LDxBlocks Workflow: From Genotype File to Haplotype Features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Overview

This vignette walks through the complete LDxBlocks workflow, from reading a
genotype file to obtaining haplotype feature matrices and diversity statistics.
The pipeline proceeds in four stages:

1. **Read** genotype data via `read_geno()` or the all-in-one `run_ldx_pipeline()`
2. **Detect** LD blocks chromosome-by-chromosome with `run_Big_LD_all_chr()`
3. **Extract** haplotypes and compute diversity with `extract_haplotypes()` and
   `compute_haplotype_diversity()`
4. **Predict** or **associate**: build a feature matrix for genomic prediction
   or run haplotype association tests

All examples use `eval = FALSE` so the vignette builds without requiring the
large input files. Adapt paths and parameters to your dataset.

---

# Reading genotype data

## Supported formats

`read_geno()` auto-detects format from the file extension:

```{r read-geno}
library(LDxBlocks)

# VCF (recommended for WGS data; auto-converted to GDS cache on first call)
be_vcf  <- read_geno("mydata.vcf.gz")

# PLINK BED (requires BEDMatrix)
be_bed  <- read_geno("mydata.bed")

# SNPRelate GDS (requires SNPRelate; fastest for large panels)
be_gds  <- read_geno("mydata.gds")

# Numeric dosage CSV (SNP × individuals, values in {0,1,2,NA})
be_csv  <- read_geno("mydata.csv", format = "numeric")

# Plain R matrix (backward-compatible)
be_mat  <- read_geno(my_matrix, snp_info = my_snp_df)
```

The `LDxBlocks_backend` object stores metadata without loading the full
genotype matrix:

```{r backend-inspect}
be_vcf
# LDxBlocks backend
#   Type    : vcf
#   Samples : 204
#   SNPs    : 2960965
#   Chr     : 1, 2, 3, ..., 12
```

Always call `close_backend()` when finished:

```{r close-backend}
close_backend(be_vcf)
```

## Chromosome label normalisation

At read time, the leading `chr` / `Chr` / `CHR` prefix is stripped from
chromosome labels. Polyploid sub-genome labels (`1A`, `2D`) are preserved
verbatim:

| Input label | Stored as |
|---|---|
| `chr1`, `Chr1`, `CHR1` | `1` |
| `Chr1A` | `1A` |
| `Chr2D` | `2D` |

---

# LD block detection

## Recommended settings

**For chip-density panels (< 100k SNPs per chromosome):**

```{r ld-chip}
blocks <- run_Big_LD_all_chr(
  geno_matrix = be,
  method      = "r2",         # standard r² — no kinship correction
  CLQcut      = 0.70,
  CLQmode     = "Density",    # exact Bron-Kerbosch clique enumeration
  subSegmSize = 1500L,
  leng        = 200L,
  MAFcut      = 0.05,
  n_threads   = 8L,
  verbose     = TRUE
)
```

**For WGS panels (2M–10M+ SNPs):**

```{r ld-wgs}
blocks <- run_Big_LD_all_chr(
  geno_matrix     = be,
  method          = "r2",
  CLQcut          = 0.80,
  CLQmode         = "Leiden",       # guaranteed connected communities, O(n log n)
  max_bp_distance = 500000L,        # skip pairs > 500 kb (70-90% of pairs)
  subSegmSize     = 500L,
  leng            = 50L,
  checkLargest    = TRUE,
  n_threads       = 8L,
  verbose         = TRUE
)
```

**For related populations (livestock, inbred lines):**

```{r ld-rv2}
blocks <- run_Big_LD_all_chr(
  geno_matrix = be,
  method      = "rV2",         # kinship-adjusted LD — requires AGHmatrix + ASRgenomics
  kin_method  = "chol",        # Cholesky whitening (default); "eigen" for near-singular GRM
  CLQcut      = 0.70,
  n_threads   = 8L
)
```

## Block table structure

```{r blocks-head}
head(blocks)
#   start  end start.rsID end.rsID start.bp   end.bp CHR length_bp
# 1     1   18      rs001    rs018     1000   103000   1    102001

summarise_blocks(blocks)
#    CHR n_blocks min_bp median_bp mean_bp max_bp total_bp_covered
# 1    1     1214   2000     82000   88940 512000        108034316
```

## Parameter auto-tuning with GWAS markers

When GWAS-significant markers are available, `tune_LD_params()` finds the
`CLQcut` value that best captures those markers within blocks:

```{r tune}
result <- tune_LD_params(
  geno_matrix    = my_geno,
  snp_info       = my_snp_info,
  gwas_df        = my_gwas,           # data.frame: Marker, CHR, POS
  prefer_perfect = TRUE,              # zero unassigned + zero forced first
  target_bp_band = c(5e4, 5e5),
  parallel       = FALSE,
  seed           = 42L
)

result$best_params       # selected parameters
result$gwas_assigned     # GWAS markers with LD_block column
result$final_blocks      # block table from best_params
```

---

# Haplotype extraction

## Phase-free (diploid allele strings)

`extract_haplotypes()` concatenates each individual's allele codes (0, 1, or 2)
for all SNPs within a block:

```{r extract}
haps <- extract_haplotypes(
  geno   = be,           # LDxBlocks_backend or plain matrix
  snp_info = be$snp_info,
  blocks = blocks,
  min_snps = 3L          # skip blocks with fewer than 3 SNPs
)

# Named list; one character vector (length n_individuals) per block
length(haps)      # number of haplotype blocks
haps[[1]][1:5]    # first 5 individuals, first block
# [1] "0120" "1100" "0210" "2100" "0021"
```

The `block_info` attribute contains block metadata:

```{r block-info}
bi <- attr(haps, "block_info")
head(bi)
#        block_id CHR start_bp  end_bp n_snps
# block_1_1000_103000   1     1000  103000     25
```

## Diversity metrics

```{r diversity}
div <- compute_haplotype_diversity(haps)

head(div[, c("block_id","CHR","n_snps","n_ind","n_haplotypes","He","Shannon","freq_dominant","sweep_flag")])

# Blocks under potential selective sweep
div[div$sweep_flag, c("block_id","CHR","start_bp","He","freq_dominant")]
```

## Haplotype feature matrix for genomic prediction

```{r feature-matrix}
feat_obj <- build_haplotype_feature_matrix(
  haplotypes     = haps,
  top_n          = 5L,         # keep top 5 most frequent alleles per block
  min_freq       = 0.01,       # discard alleles < 1% frequency
  scale_features = TRUE        # standardise columns to zero mean, unit variance
)

feat <- feat_obj$matrix        # individuals × (n_blocks × top_n)
dim(feat)                      # e.g. 204 × 52105

# Haplotype GRM for GBLUP
G_hap <- tcrossprod(feat) / ncol(feat)
```

---

# End-to-end pipeline: `run_ldx_pipeline()`

For the most common use case — VCF or BED input → blocks → haplotypes →
feature matrix — `run_ldx_pipeline()` wraps all steps in a single call:

```{r pipeline-basic}
result <- run_ldx_pipeline(
  geno_source    = "mydata.vcf.gz",
  out_dir        = "ldx_results",
  out_blocks     = "ldx_results/blocks.csv",
  out_diversity  = "ldx_results/diversity.csv",
  out_hap_matrix = "ldx_results/hap_matrix.csv",
  hap_format     = "numeric",

  maf_cut        = 0.05,
  impute         = "mean_rounded",

  CLQcut         = 0.70,
  method         = "r2",
  CLQmode        = "Leiden",
  max_bp_distance = 500000L,
  subSegmSize    = 500L,
  leng           = 50L,
  n_threads      = 8L,

  min_snps_block = 3L,
  top_n          = 5L,

  verbose = TRUE
)

result$blocks          # LD block table
result$diversity       # per-block diversity
result$haplotypes      # raw haplotype strings
result$geno_matrix     # cleaned + imputed genotype matrix
result$phase_method    # "unphased"
```

## With bigmemory caching (fast restart)

For WGS datasets, enable bigmemory caching so subsequent runs reattach the
imputed genotype matrix from disk without re-reading the raw VCF:

```{r pipeline-bigmemory}
result <- run_ldx_pipeline(
  geno_source    = "mydata.vcf.gz",
  out_dir        = "ldx_results",
  out_blocks     = "ldx_results/blocks.csv",
  out_hap_matrix = "ldx_results/hap_matrix.csv",
  CLQcut         = 0.70,
  CLQmode        = "Leiden",
  max_bp_distance = 500000L,
  n_threads      = 8L,

  use_bigmemory  = TRUE,
  bigmemory_path = "ldx_results/bm_cache",
  bigmemory_type = "char",     # 1 byte per cell; "double" for full precision

  verbose = TRUE
)
# Second run: reattaches from bm_cache instantly
```

---

# Genomic prediction

## Single-trait GBLUP

```{r gblup-single}
pred <- run_haplotype_prediction(
  geno_matrix = geno,
  snp_info    = snp_info,
  blocks      = blocks,
  blues       = setNames(my_blues_df$YLD, my_blues_df$id),
  id_col      = "id",
  blue_col    = "YLD"
)

# GEBVs
head(pred$gebv[, c("id", "YLD")])

# Most important blocks
important <- pred$block_importance[pred$block_importance$important, ]
head(important[order(-important$var_scaled), c("block_id","CHR","var_scaled")])
```

## Multi-trait GBLUP

```{r gblup-multi}
pred_mt <- run_haplotype_prediction(
  geno_matrix     = geno,
  snp_info        = snp_info,
  blocks          = blocks,
  blues           = my_blues_df,          # wide data frame
  id_col          = "id",
  blue_cols       = c("YLD", "DIS", "PHT"),
  importance_rule = "any"                 # important in at least one trait
)

# Blocks important for all three traits
all_imp <- pred_mt$block_importance[pred_mt$block_importance$important_all, ]
```

## Cross-validation

```{r cv}
cv <- cv_haplotype_prediction(
  geno_matrix = geno,
  snp_info    = snp_info,
  blocks      = blocks,
  blues       = my_blues_vec,
  k           = 5L,
  n_rep       = 3L,
  seed        = 42L
)
cv$pa_mean     # mean PA ± SD per trait across folds
```

---

# GWAS integration

## Define QTL regions

```{r qtl}
qtl <- define_qtl_regions(
  gwas_df       = my_gwas,          # data.frame: Marker, CHR, POS, P
  blocks        = blocks,
  snp_info      = snp_info,
  p_threshold   = 5e-8,
  verbose       = TRUE
)

# Pleiotropic blocks (QTL for multiple traits)
qtl[qtl$n_traits > 1, ]
```

## Integrate evidence for breeding priority

```{r priority}
priority <- integrate_gwas_haplotypes(
  qtl_regions = qtl,
  pred_result = pred_mt,
  diversity   = div
)

# Top candidates: supported by GWAS + variance + diversity
priority[priority$priority_score == 3,
         c("block_id","CHR","start_bp","priority_score")]
```

## Export candidate regions for annotation

```{r export}
export_candidate_regions(
  qtl_regions = qtl,
  format      = "bed",
  chr_prefix  = "chr",
  out_file    = "candidate_regions.bed"
)
```

---

# Haplotype association testing

## EMMAX model (pure GRM, recommended for continuous kinship)

```{r assoc-emmax}
blues_vec <- setNames(my_blues_df$YLD, my_blues_df$id)

# EMMAX: GRM random effect only, FDR-based discovery
assoc <- test_block_haplotypes(
  haplotypes = haps,
  blues      = blues_vec,
  blocks     = blocks,
  n_pcs      = 0L,          # EMMAX: GRM random effect only
  sig_metric = "p_fdr",     # Benjamini-Hochberg FDR (recommended for discovery)
  verbose    = FALSE
)

# Top blocks by omnibus significance
head(assoc$block_tests[order(assoc$block_tests$p_omnibus), ])

# Significant per-allele associations (FDR <= 0.05)
assoc$allele_tests[assoc$allele_tests$significant, ]
```

## Q+K model with simpleM correction (recommended for family-wise control)

Use `sig_metric = "p_simplem_sidak"` and `meff_scope = "chromosome"` when you
want family-wise error control that accounts for the correlation among haplotype
alleles (less conservative than raw Bonferroni). All four p-value columns are
always present in the output regardless of `sig_metric`.

```{r assoc-simplem}
assoc_sm <- test_block_haplotypes(
  haplotypes       = haps,
  blues            = blues_vec,
  blocks           = blocks,
  n_pcs            = 3L,                 # Q+K: 3 GRM-derived PCs as fixed effects
  sig_metric       = "p_simplem_sidak",  # simpleM Šidák (recommended)
  meff_scope       = "chromosome",       # separate Meff per chromosome (recommended)
  meff_percent_cut = 0.995,              # 99.5% variance threshold (Gao default)
  verbose          = FALSE
)

# Per-chromosome effective test counts
assoc_sm$meff$YLD$allele$chromosome

# All four p-value columns always present
names(assoc_sm$allele_tests)
# ... "p_wald" "p_fdr" "p_simplem" "p_simplem_sidak" "Meff" ...
```

## Automatic PC model selection (optional, recommended when lambda != 1)

When genomic control lambda is deflated (< 1, over-correction) or inflated
(> 1, under-correction), `optimize_pcs = TRUE` lets the data choose n_pcs:

```{r assoc-optimize}
assoc_opt <- test_block_haplotypes(
  haplotypes       = haps,
  blues            = blues_vec,
  blocks           = blocks,
  optimize_pcs     = TRUE,           # triggers automatic PC selection
  optimize_pcs_max = 10L,            # test k = 0..10
  optimize_method  = "bic_lambda",   # recommended: |lambda-1| + 0.01*BIC
  sig_metric       = "p_simplem_sidak",
  meff_scope       = "chromosome",
  plot             = TRUE,           # PDF plots: Manhattan, QQ, PCA, scree
  out_dir          = "my_results",
  verbose          = TRUE
)

# Model selection table: one row per k tested
assoc_opt$pc_model_selection
#  n_pcs     BIC lambda_gc   score selected
#      0  1234.1     1.021  0.0210    FALSE
#      1  1231.5     1.015  0.0150    FALSE
#      2  1230.8     1.003  0.0033    FALSE
#      3  1232.1     0.999  0.0011     TRUE

# n_pcs chosen
assoc_opt$n_pcs_used
```

## Diplotype effects (additive + dominance)

`estimate_diplotype_effects()` now supports the full correction set with a
`sig_metric` parameter, mirroring `test_block_haplotypes()`:

```{r diplotype}
dip <- estimate_diplotype_effects(
  haplotypes       = haps,
  blues            = blues_vec,
  blocks           = blocks,
  min_n_diplotype  = 3L,
  sig_metric       = "p_omnibus_simplem_sidak",  # recommended
  meff_percent_cut = 0.995,
  verbose          = FALSE
)

# All five correction columns always present in $omnibus_tests:
# p_omnibus_adj | p_omnibus_fdr | p_omnibus_simplem | p_omnibus_simplem_sidak | Meff
head(dip$omnibus_tests[, c("block_id","trait","p_omnibus","p_omnibus_fdr",
                            "p_omnibus_simplem_sidak","Meff","significant")], 5)

# Blocks showing overdominance (heterosis candidates)
dip$dominance_table[dip$dominance_table$overdominance,
                    c("block_id","allele_A","allele_B","a","d","d_over_a")]
```

## Cross-population effect concordance

`compare_block_effects()` validates whether haplotype effects discovered in one
population replicate in another, using IVW meta-analysis, Cochran Q
heterogeneity, and direction agreement:

```{r cross-pop}
# Assumes assoc_popA and assoc_popB are test_block_haplotypes() results
# from two independent populations, both run using the same block boundaries

conc <- compare_block_effects(
  assoc_pop1            = assoc_popA,
  assoc_pop2            = assoc_popB,
  pop1_name             = "PopA",
  pop2_name             = "PopB",
  blocks_pop1           = blocks_popA,
  blocks_pop2           = blocks_popB, # Pop B own block table (may differ)
  block_match           = "position",  # match by genomic IoU; match_type column reports result
  overlap_min           = 0.50,   # min IoU for 'position' match (else 'pop1_only')
  direction_threshold   = 0.75,
  boundary_overlap_warn = 0.80
)

# boundary_overlap_ratio is an OUTPUT column (automatically computed from block
# tables); boundary_overlap_warn is the INPUT threshold (default 0.80) that
# controls when boundary_warning = TRUE is set.

# Replicated blocks: directionally concordant AND Q_p > 0.05
conc$concordance[conc$concordance$replicated, ]

# Per-allele IVW detail
head(conc$shared_alleles)

print(conc)   # summary: n compared, n replicated, median I²
```

---

# Breeding decision tools

## Score individuals for haplotype stacking

```{r score}
# Get allele effects from prediction pipeline
ae <- decompose_block_effects(
  haplotypes  = haps,
  snp_info    = snp_info,
  blocks      = blocks,
  snp_effects = pred$snp_effects[["YLD"]]
)

scores <- score_favorable_haplotypes(haps, allele_effects = ae, normalize = TRUE)

# Top 10 selection candidates
head(scores[order(scores$rank), c("id","stacking_index","n_blocks_scored","rank")], 10)
```

## Parent haplotype inventory

```{r parent-inv}
top10 <- scores$id[scores$rank <= 10]

inv <- summarize_parent_haplotypes(
  haplotypes    = haps,
  candidate_ids = top10,
  allele_effects = ae
)

# Parents carrying rare alleles with positive effects
inv[inv$dosage > 0 & inv$is_rare & inv$allele_effect > 0,
    c("id","block_id","allele","dosage","allele_freq","allele_effect")]
```

---

# Between-population comparison

```{r pop-compare}
ids <- rownames(geno)

cmp <- compare_haplotype_populations(
  haplotypes  = haps,
  group1      = ids[ids %in% wild_ids],
  group2      = ids[ids %in% elite_ids],
  group1_name = "wild",
  group2_name = "elite"
)

# Divergent blocks (FST > 0.1, chi-squared p < 0.05)
cmp[cmp$divergent, c("block_id","FST","max_freq_diff","chisq_p")]
```

---

# Cross-population validation from external GWAS

When GWAS was run in an external tool (GAPIT, TASSEL, FarmCPU, PLINK), use
`compare_gwas_effects()` rather than `compare_block_effects()`.

## Pre-mapped path (recommended)

```{r gwas-conc-premapped}
# Map each population's markers to blocks first
qtl_popA <- define_qtl_regions(
  gwas_df     = my_gwas_A,
  blocks      = blocks,
  snp_info    = snp_info,
  p_threshold = 5e-8
)
qtl_popB <- define_qtl_regions(
  gwas_df     = my_gwas_B,
  blocks      = blocks,
  snp_info    = snp_info,
  p_threshold = 5e-8
)

conc_gwas <- compare_gwas_effects(
  qtl_pop1    = qtl_popA,
  qtl_pop2    = qtl_popB,
  blocks_pop1 = blocks_popA,
  blocks_pop2 = blocks_popB,
  block_match = "position",
  overlap_min = 0.50,
  pop1_name   = "PopA",
  pop2_name   = "PopB"
)

# Replicated: direction concordant AND meta_p <= 0.05
conc_gwas$concordance[conc_gwas$concordance$replicated, ]
print(conc_gwas)
```

## Raw GWAS path (convenience)

SE is derived automatically from BETA and P when absent.
`Marker` is accepted as an alias for `SNP`.

```{r gwas-conc-raw}
conc_raw <- compare_gwas_effects(
  gwas_pop1     = my_gwas_A,
  gwas_pop2     = my_gwas_B,
  blocks_pop1   = blocks,
  blocks_pop2   = blocks,
  snp_info_pop1 = snp_info,     # reused for pop2 when snp_info_pop2 = NULL
  p_threshold   = 5e-8,
  beta_col      = "BETA",       # adjust to match your GWAS output column names
  se_col        = "SE",         # NULL or absent -> derived from z-score
  p_col         = "P",
  pop1_name     = "PopA",
  pop2_name     = "PopB",
  verbose       = FALSE
)

# Extra output columns vs compare_block_effects():
#   lead_snp_pop1/pop2  — which SNP tagged the block in each population
#   lead_p_pop1/pop2    — lead SNP p-values
#   se_derived_pop1/pop2 — TRUE when SE was derived from z-score
#   both_pleiotropic    — TRUE when block is pleiotropic in both populations
head(conc_raw$concordance[, c("block_id","lead_snp_pop1","lead_snp_pop2",
                                "meta_effect","meta_p","direction_agreement",
                                "se_derived_pop1","replicated")])
```

# Session information

```{r session-info, eval=TRUE}
sessionInfo()
```
