---
title: "Large-Scale Analysis: GDS and PLINK BED Backends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Large-Scale Analysis: GDS and PLINK BED Backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  eval      = FALSE   # all chunks eval=FALSE: requires large files / optional pkgs
)
library(LDxBlocks)
```

All code chunks in this vignette have `eval = FALSE` because they require
either optional packages (SNPRelate, BEDMatrix) or large data files that are
not shipped with the package. Copy and adapt them to your own data.

---

## Why scale matters: the original Big-LD limitation

The original `Big_LD()` function (Kim et al. 2018) accepts a plain R numeric
matrix -- the entire genotype dataset must be in RAM simultaneously. For a
500-individual panel with 50,000 SNPs this works fine. For modern breeding
programmes operating with whole-genome sequencing data, the numbers become
untenable:

| Dataset | SNPs | Individuals | RAM required (naive) |
|---|---|---|---|
| Typical SNP chip | 50,000 | 5,000 | ~2 GB |
| Medium WGS panel | 1,000,000 | 500 | ~4 GB |
| Large WGS panel | 3,000,000 | 204 | ~5 GB |
| Rice mega-GWAS panel | 3,000,000 | 5,000 | ~120 GB |
| Bovine WGS | 10,000,000 | 2,000 | ~160 GB |

The 120 GB rice example is from the motivating use case for LDxBlocks:
`second_variant_calling_merged.snpsOnly.maf05.miss20.vcf.gz`, a 3M-SNP,
204-sample WGS panel where the original Big-LD would require loading the
entire genotype matrix before a single block boundary is computed.

## How LDxBlocks solves it: never-full-genome memory model

LDxBlocks enforces a strict memory contract:
**the full genotype matrix is never held in RAM at once, for any dataset size
or format.**

| Format | Memory strategy | Peak RAM for 3M x 204 panel |
|--------|----------------|---|
| VCF / HapMap | Auto-converted to GDS cache; streams per chromosome window | ~60 MB per window |
| Numeric CSV | Two-pass chunked pre-allocated reader; 50,000-row chunks | ~50 MB per chunk |
| GDS | `snpgdsGetGeno()` per window | ~60 MB per window |
| PLINK BED | `BEDMatrix` OS memory-mapping; page faults load only requested bytes | ~60 MB per window |

The peak RAM formula for file-backed backends is:

```
RAM_peak ~ n_samples x subSegmSize x 8 bytes
         = 204 x 1500 x 8 = 2.4 MB per window (3M-SNP rice panel)
         = 5000 x 1500 x 8 = 60 MB per window (5000-sample cattle panel)
```

Compare to the naive approach: `204 x 3,000,000 x 8 = 4.9 GB` just for the
matrix, before any LD computation allocates additional memory.

In all cases `run_Big_LD_all_chr()` calls `read_chunk(backend, col_idx)` to
obtain exactly the SNP window it needs, frees it with `rm()` and `gc(FALSE)`
before moving to the next window, and never accumulates more than one
chromosome in RAM at any point.

Two file-backed backends handle the on-disk streaming:

| Backend | Format | Package required |
|---------|--------|-----------------|
| `"gds"` | SNPRelate GDS | `BiocManager::install("SNPRelate")` |
| `"bed"` | PLINK binary BED | `install.packages("BEDMatrix")` |

Both backends are read-only memory-mapped: at any point only the columns
requested by the current `read_chunk()` call are in RAM.

## Memory model

Peak RAM with a file-backed backend:

$$\text{RAM}_{\text{peak}} \approx n \times w \times 8 \text{ bytes}$$

where $n$ = individuals and $w$ = `subSegmSize`. For $n = 5{,}000$ and the
default `subSegmSize = 1500`:

$$5{,}000 \times 1{,}500 \times 8 = 60 \text{ MB per window}$$

Compare to the plain-matrix path: $5{,}000 \times 10{,}000{,}000 \times 8
= 400 \text{ GB}$.

---

## GDS backend (SNPRelate)

### Installation

```{r gds-install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("SNPRelate")
```

### Converting VCF to GDS

Convert once; use for all subsequent analyses. The GDS format supports
random-access compressed storage, making repeated window reads fast:

```{r vcf-to-gds, eval=FALSE}
library(SNPRelate)
# Convert VCF to GDS once; all subsequent analyses stream from the GDS file
SNPRelate::snpgdsVCF2GDS(
  vcf.fn      = "mydata.vcf.gz",
  out.fn      = "mydata.gds",
  method      = "biallelic.only",
  snpfirstdim = FALSE
)
```

### Block detection from GDS

```{r gds-run, eval=FALSE}
be_gds <- read_geno("mydata.gds")
be_gds
# LDxBlocks backend
#   Type    : gds
#   Samples : 5000
#   SNPs    : 10241680
#   Chr     : 1, 2, ..., 29

blocks <- run_Big_LD_all_chr(
  be_gds,
  method       = "r2",
  CLQcut       = 0.70,
  leng         = 100L,      # reduce for dense WGS panels
  subSegmSize  = 1500L,     # controls peak RAM
  n_threads    = 16L,
  min_snps_chr = 100L,      # skip scaffold chromosomes
  verbose      = TRUE
)

close_backend(be_gds)
```

### What happens inside `read_chunk()` for GDS

Each call executes:

```r
dos <- SNPRelate::snpgdsGetGeno(
  gds,
  snp.id      = snp_int_ids[col_idx],
  sample.id   = sample_ids,
  snpfirstdim = FALSE,   # returns samples x SNPs
  with.id     = FALSE
)
dos <- 2L - dos  # convert REF count to ALT dosage convention
```

For `subSegmSize = 1500` and a 30-chromosome bovine genome with 10 M markers,
this is approximately $\lceil 10{,}000{,}000 / 1{,}500 \rceil \approx 6{,}700$
GDS reads total.

---

## PLINK BED backend (BEDMatrix)

### Installation

```{r bed-install, eval=FALSE}
install.packages("BEDMatrix")
```

### Block detection from BED

The `.bim` and `.fam` files must exist at the same path stem:

```{r bed-run, eval=FALSE}
be_bed <- read_geno("mydata.bed")   # reads .bim and .fam automatically
be_bed
# LDxBlocks backend
#   Type    : bed
#   Samples : 2847
#   SNPs    : 658423
#   Chr     : 1, 2, ..., 26

blocks <- run_Big_LD_all_chr(
  be_bed,
  method    = "r2",
  CLQcut    = 0.65,
  n_threads = 8L,
  verbose   = FALSE
)

close_backend(be_bed)
```

`BEDMatrix` memory-maps the `.bed` file at the OS level. Row and column
indexing triggers page faults that load only the requested bytes from disk.

---

## Parameter selection for large panels

### `subSegmSize`

The single most important memory parameter. The C++ r² kernel allocates an
$n \times w$ genotype window and a $w \times w$ LD matrix:

| Individuals (n) | subSegmSize (w) | Window RAM | LD matrix RAM |
|-----------------|-----------------|-----------|--------------|
| 1,000 | 1,500 | 12 MB | 18 MB |
| 5,000 | 1,500 | 60 MB | 18 MB |
| 5,000 | 5,000 | 200 MB | 200 MB |
| 20,000 | 1,500 | 240 MB | 18 MB |

For very large $n$, keep `subSegmSize` at 1,500 or reduce it. For moderate
$n$, increasing to 3,000–5,000 reduces GDS reads at the cost of more RAM.

### `leng`

The boundary-scan half-window in SNPs. `boundary_scan_cpp()` scans `2 * leng`
columns per candidate cut position, with cost scaling as
$O(p \times \text{leng}^2)$ per chromosome. For a 500 k-SNP chromosome with
`leng = 200` this is ~20 billion element comparisons. Reduce `leng` to 50–100
for very dense panels:

```{r leng-tradeoff, eval=FALSE}
# Dense WGS panel: reduce leng, increase subSegmSize
blocks <- run_Big_LD_all_chr(
  be_gds,
  method      = "r2",
  CLQcut      = 0.70,
  leng        = 50L,      # narrower boundary scan
  subSegmSize = 3000L,    # fewer GDS reads per chromosome
  n_threads   = 16L
)
```

### `n_threads`

OpenMP thread scaling is efficient up to approximately 8–16 threads for
typical window sizes. Benchmark on your hardware:

```{r thread-bench, eval=FALSE}
times <- sapply(c(1L, 2L, 4L, 8L, 16L), function(nt) {
  system.time(
    run_Big_LD_all_chr(be_gds, method = "r2", CLQcut = 0.7,
                        n_threads = nt, verbose = FALSE)
  )["elapsed"]
})
names(times) <- paste0(c(1,2,4,8,16), " threads")
round(times, 1)
```

---

## Haplotype analysis on large datasets

For large panels, extract haplotypes one chromosome at a time to avoid
loading the full genome matrix:

```{r hap-large, eval=FALSE}
be     <- read_geno("wgs_panel.gds")
blocks <- run_Big_LD_all_chr(be, method = "r2", CLQcut = 0.70,
                              n_threads = 16L, verbose = FALSE)

# Pass the backend directly — extract_haplotypes() detects LDxBlocks_backend
# and streams one chromosome at a time internally.
# Each chromosome is loaded, processed, rm()'d and gc(FALSE) called before
# the next chromosome. The full genome is NEVER in RAM simultaneously.
haps <- extract_haplotypes(be, be$snp_info, blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)

# Feature matrix and output (haps is a list of strings, always small in RAM)
feat <- build_haplotype_feature_matrix(haps,
                                        encoding = "additive_012",
                                        scale_features = TRUE)
write_haplotype_numeric(feat, "haplotype_matrix_dosage.csv",
                         haplotypes = haps, snp_info = be$snp_info)
write_haplotype_character(haplotypes = haps, snp_info = be$snp_info,
                           out_file = "haplotype_matrix_nucleotide.txt")
write_haplotype_diversity(div, "haplotype_diversity.csv")

close_backend(be)
```

For chromosome-specific analyses (e.g. a single chromosome only):

```{r hap-chr1, eval=FALSE}
# Filter blocks and extract by chromosome — backend still streams
haps_chr1 <- extract_haplotypes(be, be$snp_info, blocks,
                                 chr = "1", min_snps = 5)
```

---

## Troubleshooting

**GDS handle errors on Windows.** SNPRelate GDS file handles are not
fork-safe. Avoid FORK-based parallelism (`parallel::mclapply()`); use
`n_threads > 1` (OpenMP within a single process) instead.

**BEDMatrix and missing data.** BEDMatrix codes missing genotypes as `NA`.
These are correctly handled by `compute_r2_cpp()` via mean imputation per
column. No pre-imputation is required.

**"fewer than min\_snps\_chr SNPs" messages.** Many reference genomes have
hundreds of unplaced scaffolds with a handful of markers. Set
`min_snps_chr = 100` or higher to skip them automatically.

**Out-of-memory on boundary scan.** If `boundary_scan_cpp()` runs out of
RAM on very large chromosomes, reduce `leng` to 30–50. This rarely affects
block quality because large-block boundaries are evident even with a narrow
scan window.

**Very long chromosomes (> 2 M SNPs).** For panels with > 2 M SNPs per
chromosome, set `subSegmSize = 500` and `leng = 30`. The algorithm will make
more GDS reads but use far less RAM per step.

---

## References

- Lawrence M, Huber W, Pages H, et al. (2013). Software for computing and annotating genomic ranges. *PLOS Computational Biology* **9**(8):e1003118.
  <https://doi.org/10.1371/journal.pcbi.1003118>
- VanRaden PM (2008). Efficient methods to compute genomic predictions. *Journal of Dairy Science* **91**(11):4414-4423.
  <https://doi.org/10.3168/jds.2007-0980>
