---
title: "Statistical Phasing with Beagle 5.x in LDxBlocks"
author: "LDxBlocks Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Statistical Phasing with Beagle 5.x in LDxBlocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Why phase?

`extract_haplotypes()` in its default mode produces *diploid allele strings* —
the concatenation of each individual's allele codes (0, 1, or 2) at every SNP
in a block. Within a high-LD block these strings are nearly 1:1 with true
haplotype classes (Calus et al. 2008) and are sufficient for most diversity
analyses and genomic prediction tasks.

However, several analyses require knowledge of which alleles reside on the same
chromosome (gametic phase):

- **True gametic haplotype dosage**: counting how many copies of a haplotype
  allele an individual carries on separate chromosomes (0, 1, or 2) rather than
  whether the combined diploid string matches a pattern.
- **Inbreeding estimation**: fine-scale runs of homozygosity require gametic
  phasing.
- **Heterosis modelling**: identifying heterozygous diplotypes unambiguously.

LDxBlocks provides true statistical phasing via Beagle 5.x integration through
`run_ldx_pipeline(phase = TRUE)`.

---

# Requirements

## Beagle 5.x

Download the Beagle JAR file from <https://faculty.washington.edu/browning/beagle/beagle.html>:

```bash
wget https://faculty.washington.edu/browning/beagle/beagle.22Jul22.46e.jar -O beagle.jar
```

Place `beagle.jar` in your `out_dir` (the same directory you pass to
`run_ldx_pipeline()`), or supply the full path via `beagle_jar`.

## Java

Java 8 or later is required. Verify installation:

```bash
java -version
# openjdk version "11.0.21" ...
```

If `java` is not on the system PATH, supply the full path via the `java_path`
argument.

## Input format

Beagle requires VCF or VCF.gz input with an index file (`.tbi`). The
`geno_source` argument to `run_ldx_pipeline()` must therefore point to a VCF
or VCF.gz file — GDS, BED, and numeric CSV formats are not accepted when
`phase = TRUE`.

```{r input-check}
# Check: this will error early with a clear message if geno_source is not VCF
run_ldx_pipeline(
  geno_source = "mydata.bed",   # ERROR: phase = TRUE requires VCF input
  phase       = TRUE,
  ...
)
```

---

# Phasing workflow

## What happens when `phase = TRUE`

Calling `run_ldx_pipeline(phase = TRUE)` triggers the following sequence,
transparently within the pipeline:

1. **LD block detection** runs on the unphased dosage matrix (correct for LD
   computation — phasing does not change allele correlations).
2. **Beagle phasing** is called on the original VCF via `phase_with_beagle()`.
   The phased VCF is written to `out_prefix.vcf.gz` in `out_dir`.
3. **Phased VCF reading** via `read_phased_vcf()` parses the `0|1` GT fields
   into three matrices: `hap1` (gamete 1), `hap2` (gamete 2), and `dosage =
   hap1 + hap2`.
4. **Sample and SNP alignment** aligns the phased data to the cleaned/imputed
   marker set using a composite key (CHR + POS + REF + ALT + SNP ID). This is
   robust to VCF-to-VCF minor header differences.
5. **Bigmemory caching** (when `use_bigmemory = TRUE`): hap1, hap2, and dosage
   are written as three binary-backed matrices in `bigmemory_path`. On the next
   run with the same `bigmemory_path` and unchanged phased VCF, the pipeline
   reattaches from disk instantly without re-reading the VCF or re-running Beagle.
6. **Phased haplotype extraction** passes `list(hap1, hap2, dosage)` to
   `extract_haplotypes()`, which produces gametic allele strings (`g1|g2`) per
   block per individual.

## Simple example

```{r phase-simple}
library(LDxBlocks)

result <- run_ldx_pipeline(
  geno_source        = "mydata.vcf.gz",
  out_dir            = "ldx_results",
  out_blocks         = "ldx_results/blocks.csv",
  out_diversity      = "ldx_results/diversity.csv",
  out_hap_matrix     = "ldx_results/hap_matrix.csv",

  # Phasing
  phase              = TRUE,
  beagle_jar         = "ldx_results/beagle.jar",  # or NULL to auto-find
  beagle_threads     = 8L,
  beagle_java_mem_gb = 16L,    # -Xmx16g; NULL = JVM default (~256 MB — too small for WGS)
  beagle_seed        = 42L,    # integer seed for reproducibility

  # Block detection
  CLQcut         = 0.70,
  CLQmode        = "Leiden",
  max_bp_distance = 500000L,
  subSegmSize    = 500L,
  leng           = 50L,
  n_threads      = 8L,

  # Haplotype extraction
  min_snps_block = 3L,
  top_n          = 5L,

  # Bigmemory caching (fast restart)
  use_bigmemory  = TRUE,
  bigmemory_path = "ldx_results/bm_cache",

  verbose = TRUE
)

result$phase_method   # "beagle"
result$phased_vcf     # "ldx_results/mydata_phased.vcf.gz"
```

## Verifying phased output

```{r phase-verify}
# Phase method
result$phase_method
# [1] "beagle"

# Path to phased VCF
result$phased_vcf
# [1] "ldx_results/mydata_phased.vcf.gz"

# Haplotype strings now contain "|" separating gamete 1 and gamete 2
result$haplotypes[[1]][1:4]
# [1] "0010|1010"  "1100|0100"  "0000|0110"  "1010|0000"

# Diversity table: phased = TRUE
head(result$diversity[, c("block_id","He","phased")])
```

---

# Bigmemory caching for phased data

For WGS datasets, reading a phased VCF with 3M SNPs and 204 individuals through
the R `read_phased_vcf()` parser takes several minutes. The bigmemory caching
mechanism eliminates this cost on all runs after the first.

## How the cache works

When `use_bigmemory = TRUE` and `phase = TRUE`, after reading the phased VCF,
three binary-backed matrices are written to `bigmemory_path`:

| File | Content | Type |
|------|---------|------|
| `ldxblocks_bm_phased_hap1.bin` | Gamete 1 alleles (individuals × SNPs) | `char` (0/1) |
| `ldxblocks_bm_phased_hap2.bin` | Gamete 2 alleles (individuals × SNPs) | `char` (0/1) |
| `ldxblocks_bm_phased_dos.bin` | Dosage = hap1 + hap2 (individuals × SNPs) | `char` (0/1/2) |

These are accompanied by `.desc` descriptor files, `_snpinfo.rds`,
`_sampleids.rds`, and `_params.rds` (the fingerprint).

**Fingerprint:** VCF path + mtime + target SNP count + sample count. If all
cache files exist and the fingerprint matches, the VCF is not re-read.

**Disk space:** For 204 individuals × 2.96M SNPs at 1 byte per cell, each
matrix requires ~600 MB. Total for three matrices: ~1.8 GB.

## Forcing a cache rebuild

To invalidate the cache (e.g., after re-running Beagle with different
parameters), delete the `ldxblocks_bm_phased_*` files from `bigmemory_path`:

```{r cache-invalidate}
# From R
files <- list.files("ldx_results/bm_cache", pattern = "ldxblocks_bm_phased",
                    full.names = TRUE)
file.remove(files)

# Then rerun — will re-read the phased VCF and rebuild the cache
result <- run_ldx_pipeline(..., phase = TRUE, use_bigmemory = TRUE, ...)
```

---

# Using `phase_with_beagle()` directly

If you need finer control over the phasing step (e.g., chromosome-by-chromosome
phasing, or phasing with a reference panel), you can call `phase_with_beagle()`
directly:

```{r beagle-direct}
phased_vcf <- phase_with_beagle(
  input_vcf   = "mydata.vcf.gz",
  out_prefix  = "ldx_results/mydata_phased",
  beagle_jar  = "ldx_results/beagle.jar",
  java_path   = "java",
  java_mem_gb = 16L,            # -Xmx16g JVM heap
  nthreads    = 8L,
  ref_panel   = NULL,           # phased reference VCF (optional)
  map_file    = NULL,           # genetic map for improved accuracy (optional)
  chrom       = "1",            # restrict to chromosome 1 (optional)
  seed        = 42L,
  burnin      = 6L,             # Beagle default: 6
  iterations  = 12L,            # Beagle default: 12
  window      = 25.0,           # window size in cM (Beagle default: 25)
  overlap     = 2.5,            # window overlap in cM (Beagle default: 2.5)
  verbose     = TRUE
)

phased_vcf
# [1] "ldx_results/mydata_phased.vcf.gz"
```

The Beagle log is written to `ldx_results/mydata_phased.log`.

## Using a reference panel

A phased reference VCF substantially improves phasing accuracy, especially for
rare variants. Reference panels are available from the 1000 Genomes Project,
HapMap, or species-specific databases:

```{r beagle-ref}
phased_vcf <- phase_with_beagle(
  input_vcf   = "mydata.vcf.gz",
  out_prefix  = "ldx_results/mydata_phased",
  beagle_jar  = "ldx_results/beagle.jar",
  ref_panel   = "reference_panel.vcf.gz",   # must be phased (0|1 GT fields)
  java_mem_gb = 32L,
  nthreads    = 8L
)
```

## Chromosome-by-chromosome phasing

For very large datasets, phasing one chromosome at a time reduces peak memory:

```{r beagle-chr}
chrs <- as.character(1:12)

phased_vcfs <- lapply(chrs, function(chr) {
  phase_with_beagle(
    input_vcf   = "mydata.vcf.gz",
    out_prefix  = sprintf("ldx_results/mydata_phased_chr%s", chr),
    beagle_jar  = "ldx_results/beagle.jar",
    chrom       = chr,
    java_mem_gb = 8L,
    nthreads    = 8L,
    seed        = 42L
  )
})
```

---

# Reading a pre-phased VCF

If your data are already phased (from Beagle, SHAPEIT, or another tool),
use `read_phased_vcf()` directly:

```{r read-phased}
gp <- read_phased_vcf("mydata_phased.vcf.gz", min_maf = 0.01, verbose = TRUE)

# Returns a list:
names(gp)
# [1] "hap1"       "hap2"       "dosage"     "snp_info"   "sample_ids" "phased"     "phase_method"

dim(gp$hap1)     # SNPs × individuals
dim(gp$hap2)     # SNPs × individuals
dim(gp$dosage)   # SNPs × individuals (= hap1 + hap2)

gp$phased        # TRUE
gp$phase_method  # "vcf_phased"
```

Pass the resulting list directly to `extract_haplotypes()`:

```{r extract-phased}
haps <- extract_haplotypes(
  geno     = gp,           # list with hap1 and hap2 elements
  snp_info = gp$snp_info,
  blocks   = blocks,
  min_snps = 3L
)

# Haplotype strings contain gametic phase
haps[[1]][1:3]
# [1] "0010|1010"  "1100|0100"  "0000|0110"
```

---

# Phased haplotype feature matrix

When gametic phase is available, `build_haplotype_feature_matrix()` produces
true allele copy number (0, 1, or 2):

```{r phased-features}
feat_phased <- build_haplotype_feature_matrix(
  haplotypes     = haps,          # from phased extraction
  top_n          = 5L,
  scale_features = TRUE
)$matrix

# Each column encodes: 0 = neither gamete, 1 = heterozygous, 2 = homozygous
# This is true allele copy number — not present/absent
range(feat_phased)   # [0, 2] (before scaling)
```

Compare to unphased:

```{r unphased-features}
feat_unphased <- build_haplotype_feature_matrix(
  haplotypes     = haps_unphased,   # from dosage-string extraction
  top_n          = 5L,
  scale_features = FALSE
)$matrix

range(feat_unphased)   # [0, 1] — presence/absence only
```

---

# Diversity metrics with phased data

The `compute_haplotype_diversity()` function automatically detects phased input
and sets the `phased` column to `TRUE`. He correction uses the number of
gametes (2N) rather than individuals (N):

```{r phased-diversity}
div_phased <- compute_haplotype_diversity(haps)

# phased column
head(div_phased[, c("block_id","n_ind","He","phased")])
#         block_id n_ind      He phased
# block_1_1000_103000   204  0.891   TRUE
```

---

# Haplotype association with phased data

`test_block_haplotypes()` uses phased dosage (0/1/2) when phased haplotype
strings are passed, giving unambiguous allele copy-number encoding for the
association model. Use `sig_metric = "p_simplem_sidak"` for family-wise error
control with correlated haplotype predictors (recommended):

```{r assoc-phased}
blues_vec <- setNames(my_blues$YLD, my_blues$id)

assoc_phased <- test_block_haplotypes(
  haplotypes       = haps,              # phased (contains "|")
  blues            = blues_vec,
  blocks           = blocks,
  n_pcs            = 3L,               # Q+K: GRM PCs as fixed effects
  sig_metric       = "p_simplem_sidak", # simpleM Šidák correction
  meff_scope       = "chromosome",      # separate Meff per chromosome
  meff_percent_cut = 0.995,
  verbose          = FALSE
)

# Effect estimates reflect true additive copy-number dosage (0/1/2)
head(assoc_phased$allele_tests[order(assoc_phased$allele_tests$p_wald),
                                c("block_id","allele","effect","SE",
                                  "p_wald","p_simplem","p_simplem_sidak","Meff")])

# All four p-value columns are always present in the output
# (p_wald, p_fdr, p_simplem, p_simplem_sidak) regardless of sig_metric
```

---

# Troubleshooting

## `beagle.jar` not found

```
Error: beagle.jar not found at ldx_results/beagle.jar.
  Download from https://faculty.washington.edu/browning/beagle/beagle.html
  and place in out_dir, or supply the full path via beagle_jar.
```

**Fix:** Download `beagle.jar` and either place it in `out_dir` or pass its
full path via `beagle_jar = "/path/to/beagle.jar"`.

Note: `Sys.which("beagle.jar")` typically returns `""` — Java `.jar` files are
not executables and are not found by `which`. Always supply the path explicitly.

## Out-of-memory error in Beagle

```
Exception in thread "main" java.lang.OutOfMemoryError: Java heap space
```

**Fix:** Increase `beagle_java_mem_gb`. For WGS data (3M SNPs), 16–32 GB is
recommended:

```{r oom-fix}
result <- run_ldx_pipeline(
  ...,
  beagle_java_mem_gb = 32L,   # -Xmx32g
  ...
)
```

## VCF index missing

```
Error: .tbi index not found for mydata.vcf.gz.
  Run: tabix -p vcf mydata.vcf.gz
```

**Fix:** Index the VCF with tabix (requires htslib):

```bash
tabix -p vcf mydata.vcf.gz
```

## Phased SNP count mismatch after alignment

```
Error: Phased VCF missing 1234 SNP(s) from the cleaned/imputed marker set
  after CHR+POS+REF+ALT+SNP matching.
```

**Fix:** This occurs when the phased VCF's marker set differs from the cleaned
set (e.g., Beagle removed some SNPs). Check `result$snp_info_filtered` against
the SNPs in the phased VCF. Consider using `clean_malformed = FALSE` in
`run_ldx_pipeline()` if malformed-record removal is removing too many markers.

## Cache not reattaching

If the phased VCF has been modified or re-generated, the fingerprint will
not match and the pipeline will rebuild the cache. This is the expected
behaviour. If you want to force a rebuild explicitly:

```{r force-rebuild}
# Delete cache files
unlink(list.files("ldx_results/bm_cache",
                   pattern = "ldxblocks_bm_phased", full.names = TRUE))
```

---

# Summary of phasing parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `phase` | logical | `FALSE` | Enable Beagle phasing |
| `beagle_jar` | character | `file.path(out_dir, "beagle.jar")` | Path to `beagle.jar` |
| `beagle_threads` | integer | `1L` | Beagle CPU threads |
| `java_path` | character | `"java"` | Path to Java executable |
| `beagle_java_mem_gb` | integer or NULL | `NULL` | JVM heap size (GB). `NULL` = JVM default. WGS: use 16–32. |
| `beagle_seed` | integer or NULL | `NULL` | Random seed for reproducibility |
| `beagle_ref_panel` | character or NULL | `NULL` | Phased reference VCF (improves rare-variant phasing) |
| `beagle_map_file` | character or NULL | `NULL` | Genetic map file for improved recombination modelling |
| `beagle_chrom` | character or NULL | `NULL` | Restrict phasing to one chromosome |
| `beagle_burnin` | integer or NULL | `NULL` | Beagle burnin iterations (default: 6) |
| `beagle_iterations` | integer or NULL | `NULL` | Beagle phasing iterations (default: 12) |
| `beagle_window` | numeric or NULL | `NULL` | Window size in cM (default: 25) |
| `beagle_overlap` | numeric or NULL | `NULL` | Window overlap in cM (default: 2.5) |
| `beagle_args` | character | `""` | Additional Beagle command-line arguments |

---

# Session information

```{r session-info, eval=TRUE}
sessionInfo()
```
