---
title: "LD Metrics: Standard r² and Kinship-Adjusted rV²"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LD Metrics: Standard r² and Kinship-Adjusted rV²}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message   = FALSE,
  warning   = FALSE
)
library(LDxBlocks)
data(ldx_geno,     package = "LDxBlocks")
data(ldx_snp_info, package = "LDxBlocks")
```

## 1. The kinship problem in LD estimation

The standard squared Pearson correlation between two SNP dosage vectors
$g_j$ and $g_k$ is:

$$r^2_{jk} = \left[\frac{\mathrm{Cov}(g_j, g_k)}
  {\sqrt{\mathrm{Var}(g_j)\,\mathrm{Var}(g_k)}}\right]^2$$

This estimator assumes the $n$ individuals are exchangeable — drawn
independently from the same population. In structured or related populations
this assumption fails. Two individuals who share a recent common ancestor
carry correlated alleles at *all* loci, not just loci in true LD. The sample
covariance therefore picks up both genuine recombination-based LD *and*
kinship-induced allele sharing, inflating $r^2$ relative to the true
population value.

The consequence for block detection is that blocks are drawn too broadly:
regions that are genuinely independent appear correlated in the sample, the
clique graph has more edges than it should, and singletons are absorbed into
large blocks.

## 2. The kinship correction: rV²

Mangin et al. (2012) proposed replacing $r^2$ with the **kinship-adjusted
squared correlation** $rV^2$ specifically to correct LD estimates in
structured or related populations:

$$rV^2_{jk} = \left[\mathrm{Cor}(V^{-1/2} g_j,\; V^{-1/2} g_k)\right]^2$$

where $V = ZZ^\top / (2\sum_j p_j q_j)$ is the VanRaden (2008) GRM
($Z_{ij} = G_{ij} - 2p_j$, $G$ = raw dosage matrix, $p_j$ = allele
frequency at SNP $j$), called $K$ by Mangin et al. (2012) and $G$
by VanRaden (2008) himself. In LDxBlocks the symbol $V$ is used to
avoid collision with $G$ the raw genotype matrix.
$V^{-1/2}$ is its inverse square root. Left-multiplying the mean-centred
genotype matrix $\tilde{G}$ by $V^{-1/2}$ removes the kinship-induced
correlation between individuals, so the resulting correlation measures
only recombination-based LD.

## 3. Decision table: r² versus rV²

```{r decision-table, echo=FALSE}
dt <- data.frame(
  Criterion = c("Population type", "Computational cost",
                "RAM requirement", "Marker scale",
                "Block accuracy in related pops",
                "External dependencies"),
  `r² (method = 'r2')` = c(
    "Random mating, unrelated, weakly structured",
    "O(np) prep + O(p²) per window via C++",
    "Proportional to one window",
    "Scales to 10 M+ markers",
    "Slightly inflated — blocks too broad",
    "None (always installed)"
  ),
  `rV² (method = 'rV2')` = c(
    "Livestock, inbred lines, family-based cohorts",
    "O(n²p) GRM + O(n³) Cholesky + O(np) whitening",
    "n×n GRM + n×n whitening factor per chromosome",
    "Practical to ~200 k markers per chromosome",
    "Correct for structured populations",
    "AGHmatrix, ASRgenomics (optional Suggests)"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(dt, caption = "r² versus rV²: decision table")
```

## 4. Computing rV² in LDxBlocks

The whitening is handled by `prepare_geno()`. The subsequent correlation
computation uses the same C++ kernel as the standard r² path.

```{r prepare-demo}
# r² path: just mean-centre
prep_r2 <- prepare_geno(ldx_geno[, 1:30], method = "r2")
str(prep_r2)   # adj_geno = centred matrix; V_inv_sqrt = NULL

# rV² path — requires AGHmatrix + ASRgenomics (optional Suggests)
# prep_rv2 <- prepare_geno(ldx_geno[, 1:30], method = "rV2")
# str(prep_rv2)  # adj_geno = whitened matrix; V_inv_sqrt = n x n matrix
```

### 4.1 The VanRaden GRM

The kinship matrix $V$ is the additive genomic relationship matrix of
VanRaden (2008), computed by `AGHmatrix::Gmatrix()` inside `prepare_geno()`:

$$V = \frac{ZZ^\top}{2\sum_j p_j q_j}$$

where $Z$ is the $n \times m$ frequency-centred genotype matrix with
elements $z_{ij} = g_{ij} - 2p_j$ ($p_j$ = frequency of allele $A_2$
at SNP $j$, $q_j = 1 - p_j$). The coefficients at SNP $j$ are:
$0 - 2p_j$ (homozygous $A_1A_1$), $1 - 2p_j$ (heterozygous $A_1A_2$),
$2 - 2p_j$ (homozygous $A_2A_2$).
The scalar denominator $2\sum_j p_j q_j$ scales $V$ so that the
diagonal elements average to approximately 1 (resembles the numerator
relationship matrix for an outbred population).

### 4.2 The whitening factor

`get_V_inv_sqrt()` computes $A = V^{-1/2}$ such that $A V A^\top = I$.

**Cholesky** (`kin_method = "chol"`, default): $A = R^{-1}$ where
$V = R^\top R$. Fast and numerically stable.

**Eigendecomposition** (`kin_method = "eigen"`):
$A = Q \Lambda^{-1/2} Q^\top$. Eigenvalues floored at $10^{-6}$ for
stability. Symmetric result; preferable when $V$ is near-singular.

```{r whitening-demo}
set.seed(1)
G_small <- ldx_geno[, 1:30]
Gc      <- scale(G_small, center = TRUE, scale = FALSE)
V_demo  <- tcrossprod(Gc) / 30 + diag(0.05, nrow(Gc))  # toy PD matrix

A_chol <- get_V_inv_sqrt(V_demo, method = "chol")
A_eig  <- get_V_inv_sqrt(V_demo, method = "eigen")

# Verify: A V A' ≈ I
max(abs(A_chol %*% V_demo %*% t(A_chol) - diag(120)))
max(abs(A_eig  %*% V_demo %*% t(A_eig)  - diag(120)))
```

## 5. Comparing r² and rV² on example data

```{r compare-matrices}
idx_25 <- 1:25
r2_mat <- compute_r2(ldx_geno[, idx_25])

# rV² with a toy kinship (normally produced by prepare_geno)
Gc_25  <- scale(ldx_geno[, idx_25], center = TRUE, scale = FALSE)
V_toy  <- tcrossprod(Gc_25) / 25 + diag(0.1, 120)
A_toy  <- get_V_inv_sqrt(V_toy)
X_whit <- A_toy %*% Gc_25
rv2_mat <- compute_rV2(X_whit)

cat("Mean r² :", round(mean(r2_mat[upper.tri(r2_mat)]),  4), "\n")
cat("Mean rV²:", round(mean(rv2_mat[upper.tri(rv2_mat)]), 4), "\n")
```

```{r compare-image, fig.cap="r² (left) vs rV² (right) for the first 25 SNPs on chr1", fig.width=9}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
image(r2_mat,   main = "Standard r²",
      col = hcl.colors(20, "YlOrRd", rev = TRUE),
      xlab = "SNP", ylab = "SNP")
image(rv2_mat,  main = "Kinship-adjusted rV²",
      col = hcl.colors(20, "YlOrRd", rev = TRUE),
      xlab = "SNP", ylab = "SNP")
par(mfrow = c(1, 1))
```

## 6. When r² gives wrong blocks

In a livestock panel where sires appear repeatedly through progeny, the
kinship-induced correlation between half-sibs inflates $r^2$ for all SNP
pairs in the same family cluster. A pair of SNPs on different chromosomes
can show $r^2 > 0.5$ purely because both correlate with family membership.
The whitened matrix removes this signal: after left-multiplication by
$V^{-1/2}$, the effective covariance between individuals is identity, so
only genuine gametic LD contributes to $rV^2$.

In practice, rV² blocks are 10–30% smaller and more precisely delimited in
related populations. The effect is greatest in:

- Livestock populations with many half-sib families
- Inbred plant populations (F2, RILs, near-isogenic lines)
- Multi-family human cohorts from isolated populations

## 7. When r² is preferable

For large human biobank panels (n > 10,000, p > 1,000,000), computing and
inverting the n × n GRM is computationally infeasible. The standard r²
inflates LD modestly but the effect on block boundaries is small in
approximately random-mating populations, and the 50× speed advantage of the
pure C++ path dominates.

## 8. The C++ LD kernel

Both `compute_r2()` and `compute_rV2()` call `compute_r2_cpp()`.
The implementation in `src/ld_core.cpp`:

1. **NA imputation** — missing values replaced by column mean in one O(np) pass.
2. **Standardisation** — columns centred and scaled to zero mean, unit variance.
3. **Correlation matrix** — $Z^\top Z / (n-1)$ where $Z$ is the standardised
   matrix. For $p = 1{,}500$ and $n = 500$ this is a 1,500 × 1,500 DGEMM
   operation, with the upper-triangular loop parallelised by OpenMP.
4. **Squaring and clamping** — $r^2 = r \times r$, clamped to $[0, 1]$.
5. **Symmetrisation** and optional rounding (`digits`).

```{r cpp-direct, eval=FALSE}
G_30 <- ldx_geno[, 1:30]

r2_raw   <- compute_r2_cpp(G_30, digits = -1L, n_threads = 1L)  # no rounding
r2_round <- compute_r2_cpp(G_30, digits =  6L, n_threads = 1L)  # 6 dp

max(abs(r2_raw - r2_round))   # < 5e-7
```

```{r speed}
G_bench  <- ldx_geno[, 1:50]
Gc_bench <- scale(G_bench, center = TRUE, scale = FALSE)

t_r2  <- system.time(for(i in 1:20) compute_r2(G_bench))
t_rv2 <- system.time(for(i in 1:20) compute_rV2(Gc_bench))

cat("compute_r2  (50 SNPs, 20 reps):", round(t_r2["elapsed"],  3), "s\n")
cat("compute_rV2 (50 SNPs, 20 reps):", round(t_rv2["elapsed"], 3), "s\n")
# Both use the same C++ kernel — times should be identical
# The cost of rV² is entirely in prepare_geno() (GRM + Cholesky)
```

## 9. Switching between metrics

Switching is a single argument change. No other code needs to change:

```{r switch, eval=FALSE}
blocks_r2  <- run_Big_LD_all_chr(be, method = "r2",  CLQcut = 0.70)
blocks_rv2 <- run_Big_LD_all_chr(be, method = "rV2", CLQcut = 0.70,
                                  kin_method = "chol")
```

## 10. LD decay and the parametric threshold

The parametric threshold in `compute_ld_decay(r2_threshold = "parametric")`
estimates the 95th percentile of r² between markers on different chromosomes
(unlinked markers). This is conceptually identical to the kinship inflation
demonstrated in Sections 5 and 6 above: unlinked markers should have r² = 0
in an ideal population; any inflation above zero is caused by kinship/structure.

```{r ld-decay-demo, eval=FALSE}
decay <- compute_ld_decay(
  ldx_geno, ldx_snp_info,
  r2_threshold = "both",    # parametric + fixed 0.1
  n_pairs = 2000L, verbose = FALSE
)
# High critical_r2_param -> structured population -> use method = "rV2"
cat("Parametric threshold:", round(decay$critical_r2_param, 4), "\n")
cat("Standard 0.1 threshold: fixed\n")
cat("Decay distances (kb):\n")
print(decay$decay_dist[, c("CHR", "decay_dist_kb", "censored")])
```

A high parametric threshold (e.g. > 0.05) is direct evidence that
`method = "rV2"` should be used for block detection -- the same kinship
inflation that this vignette demonstrates algebraically is visible in the
distribution of unlinked-marker r² values.

## 11. References

- Kim S-A, Cho C-S, Kim S-R, Bull SB, Yoo Y-J (2018). A new haplotype block detection method for dense genome sequencing data based on interval graph modeling and dynamic programming. *Bioinformatics* **34**(4):588-596.
  <https://doi.org/10.1093/bioinformatics/btx609>
- Mangin B, Siberchicot A, Nicolas S, Doligez A, This P, Cierco-Ayrolles C (2012). Novel measures of linkage disequilibrium that correct the bias due to population structure and relatedness. *Heredity* **108**(3):285-291.
  <https://doi.org/10.1038/hdy.2011.73>
- VanRaden PM (2008). Efficient methods to compute genomic predictions. *Journal of Dairy Science* **91**(11):4414-4423.
  <https://doi.org/10.3168/jds.2007-0980>
