---
title: "Large-Scale Analysis: GDS and PLINK BED Backends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Large-Scale Analysis: GDS and PLINK BED Backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  eval      = FALSE   # all chunks eval=FALSE: requires large files / optional pkgs
)
library(LDxBlocks)
```

All code chunks in this vignette have `eval = FALSE` because they require
either optional packages (SeqArray, BEDMatrix) or large data files that are
not shipped with the package. Copy and adapt them to your own data.

---

## The scale problem

A 10 M-marker, 5,000-individual genotype matrix occupies roughly 200 GB as a
dense R `double` matrix. Loading it into RAM before block detection is
impractical on any workstation. LDxBlocks solves this through the
`LDxBlocks_backend` abstraction: `run_Big_LD_all_chr()` calls
`read_chunk(backend, col_idx)` to obtain exactly the SNP window it needs,
leaving the rest of the genome on disk.

Two file-backed backends handle this streaming access:

| Backend | Format | Package required |
|---------|--------|-----------------|
| `"gds"` | SeqArray GDS | `BiocManager::install("SeqArray")` |
| `"bed"` | PLINK binary BED | `install.packages("BEDMatrix")` |

Both backends are read-only memory-mapped: at any point only the columns
requested by the current `read_chunk()` call are in RAM.

## Memory model

Peak RAM with a file-backed backend:

$$\text{RAM}_{\text{peak}} \approx n \times w \times 8 \text{ bytes}$$

where $n$ = individuals and $w$ = `subSegmSize`. For $n = 5{,}000$ and the
default `subSegmSize = 1500`:

$$5{,}000 \times 1{,}500 \times 8 = 60 \text{ MB per window}$$

Compare to the plain-matrix path: $5{,}000 \times 10{,}000{,}000 \times 8
= 400 \text{ GB}$.

---

## GDS backend (SeqArray)

### Installation

```{r gds-install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("SeqArray")
```

### Converting VCF to GDS

Convert once; use for all subsequent analyses. The GDS format supports
random-access compressed storage, making repeated window reads fast:

```{r vcf-to-gds, eval=FALSE}
library(SeqArray)
seqVCF2GDS(
  vcf.fn         = "mydata.vcf.gz",
  out.fn         = "mydata.gds",
  storage.option = "ZIP_RA",   # random-access LZ4 compression
  parallel       = 4L
)
```

### Block detection from GDS

```{r gds-run, eval=FALSE}
be_gds <- read_geno("mydata.gds")
be_gds
# LDxBlocks backend
#   Type    : gds
#   Samples : 5000
#   SNPs    : 10241680
#   Chr     : 1, 2, ..., 29

blocks <- run_Big_LD_all_chr(
  be_gds,
  method       = "r2",
  CLQcut       = 0.70,
  leng         = 100L,      # reduce for dense WGS panels
  subSegmSize  = 1500L,     # controls peak RAM
  n_threads    = 16L,
  min_snps_chr = 100L,      # skip scaffold chromosomes
  verbose      = TRUE
)

close_backend(be_gds)
```

### What happens inside `read_chunk()` for GDS

Each call executes:

```r
SeqArray::seqSetFilter(gds, variant.id = var_ids[col_idx], verbose = FALSE)
dos <- t(SeqArray::seqGetData(gds, "$dosage"))
SeqArray::seqResetFilter(gds, verbose = FALSE)
```

For `subSegmSize = 1500` and a 30-chromosome bovine genome with 10 M markers,
this is approximately $\lceil 10{,}000{,}000 / 1{,}500 \rceil \approx 6{,}700$
GDS reads total.

---

## PLINK BED backend (BEDMatrix)

### Installation

```{r bed-install, eval=FALSE}
install.packages("BEDMatrix")
```

### Block detection from BED

The `.bim` and `.fam` files must exist at the same path stem:

```{r bed-run, eval=FALSE}
be_bed <- read_geno("mydata.bed")   # reads .bim and .fam automatically
be_bed
# LDxBlocks backend
#   Type    : bed
#   Samples : 2847
#   SNPs    : 658423
#   Chr     : 1, 2, ..., 26

blocks <- run_Big_LD_all_chr(
  be_bed,
  method    = "r2",
  CLQcut    = 0.65,
  n_threads = 8L,
  verbose   = FALSE
)

close_backend(be_bed)
```

`BEDMatrix` memory-maps the `.bed` file at the OS level. Row and column
indexing triggers page faults that load only the requested bytes from disk.

---

## Parameter selection for large panels

### `subSegmSize`

The single most important memory parameter. The C++ r² kernel allocates an
$n \times w$ genotype window and a $w \times w$ LD matrix:

| Individuals (n) | subSegmSize (w) | Window RAM | LD matrix RAM |
|-----------------|-----------------|-----------|--------------|
| 1,000 | 1,500 | 12 MB | 18 MB |
| 5,000 | 1,500 | 60 MB | 18 MB |
| 5,000 | 5,000 | 200 MB | 200 MB |
| 20,000 | 1,500 | 240 MB | 18 MB |

For very large $n$, keep `subSegmSize` at 1,500 or reduce it. For moderate
$n$, increasing to 3,000–5,000 reduces GDS reads at the cost of more RAM.

### `leng`

The boundary-scan half-window in SNPs. `boundary_scan_cpp()` scans `2 * leng`
columns per candidate cut position, with cost scaling as
$O(p \times \text{leng}^2)$ per chromosome. For a 500 k-SNP chromosome with
`leng = 200` this is ~20 billion element comparisons. Reduce `leng` to 50–100
for very dense panels:

```{r leng-tradeoff, eval=FALSE}
# Dense WGS panel: reduce leng, increase subSegmSize
blocks <- run_Big_LD_all_chr(
  be_gds,
  method      = "r2",
  CLQcut      = 0.70,
  leng        = 50L,      # narrower boundary scan
  subSegmSize = 3000L,    # fewer GDS reads per chromosome
  n_threads   = 16L
)
```

### `n_threads`

OpenMP thread scaling is efficient up to approximately 8–16 threads for
typical window sizes. Benchmark on your hardware:

```{r thread-bench, eval=FALSE}
times <- sapply(c(1L, 2L, 4L, 8L, 16L), function(nt) {
  system.time(
    run_Big_LD_all_chr(be_gds, method = "r2", CLQcut = 0.7,
                        n_threads = nt, verbose = FALSE)
  )["elapsed"]
})
names(times) <- paste0(c(1,2,4,8,16), " threads")
round(times, 1)
```

---

## Haplotype analysis on large datasets

For large panels, extract haplotypes one chromosome at a time to avoid
loading the full genome matrix:

```{r hap-large, eval=FALSE}
be      <- read_geno("wgs_panel.gds")
blocks  <- run_Big_LD_all_chr(be, method = "r2", CLQcut = 0.70,
                               n_threads = 16L, verbose = FALSE)

# Process chromosome 1 only
chr1_idx    <- which(be$snp_info$CHR == "1")
chr1_geno   <- read_chunk(be, chr1_idx)          # load chr1 only
chr1_blocks <- blocks[blocks$CHR == "1", ]
chr1_info   <- be$snp_info[chr1_idx, ]

haps <- extract_haplotypes(chr1_geno, chr1_info, chr1_blocks, min_snps = 5)
div  <- compute_haplotype_diversity(haps)
feat <- build_haplotype_feature_matrix(haps, top_n = 5, scale_features = TRUE)

close_backend(be)
```

---

## Troubleshooting

**GDS handle errors on Windows.** SeqArray GDS file handles are not
fork-safe. Avoid FORK-based parallelism (`parallel::mclapply()`); use
`n_threads > 1` (OpenMP within a single process) instead.

**BEDMatrix and missing data.** BEDMatrix codes missing genotypes as `NA`.
These are correctly handled by `compute_r2_cpp()` via mean imputation per
column. No pre-imputation is required.

**"fewer than min\_snps\_chr SNPs" messages.** Many reference genomes have
hundreds of unplaced scaffolds with a handful of markers. Set
`min_snps_chr = 100` or higher to skip them automatically.

**Out-of-memory on boundary scan.** If `boundary_scan_cpp()` runs out of
RAM on very large chromosomes, reduce `leng` to 30–50. This rarely affects
block quality because large-block boundaries are evident even with a narrow
scan window.

**Very long chromosomes (> 2 M SNPs).** For panels with > 2 M SNPs per
chromosome, set `subSegmSize = 500` and `leng = 30`. The algorithm will make
more GDS reads but use far less RAM per step.

---

## References

- Lawrence M et al. (2013) *PLOS Comput. Biol.* **9**(8):e1003118.
  <https://doi.org/10.1371/journal.pcbi.1003118>
- VanRaden PM (2008) *J. Dairy Sci.* **91**(11):4414–4423.
  <https://doi.org/10.3168/jds.2007-0980>
